---
title: "Multivariate Forecasts"
author: "73-423 Forecasting for Economics and Business"
output:
  html_document:
    code_folding: hide
---
    
## Today 

- Multivariate forecasts
- Multi-period forecasts
- Vector Autoregression (VAR) methods

## Multivariate Forecasts

- Previous problems we have seen in class involve forecasting a single number $y_{T+h}$
- Often need to forecast more than a single object
    - Need to know multiple aspects of future situationto make policy or decisions
        - Ideal monetery policy depends on future state of both inflation and unemployment
    - May need to make multiple related decisions
        - Business may care about performance of each of its many establishments
- Even when forecasting a single target, can help to use multiple variables
    - May be used directly to improve predictions
    - May be used indirectly to construct a better univariate forecast from a multivariate one
    
## Multivariate Forecasting Setup

- Similar to univariate case, except we care about $m$ different outcomes 
- Formally, data $\mathbf{y}_t=(y_{1t},y_{2t},\ldots,y_{mt})^{\prime}$ is an m-dimensional **vector** in $\mathbf{Y}=\mathbb{R}^m$
- Data available at time t is to build a forecast is $\mathcal{Z}_t=\{\mathcal{Y}_t,\mathcal{X}_t\}\in\underset{t}{\times}\mathbf{Z}=\underset{t}{\times}(\mathbf{Y}\times\mathbf{X})$
    - $\mathcal{Y}_t=\{\mathbf{y}_s\}_{s=1}^{t}$ are past observations of the series to be forecast
    - $\mathcal{X}_t=\{\mathbf{x}_s\}_{s=1}^{t}$ are past observations of other series which are not the target of forecast, but may be used for forecasting
- Produce a multivariate forecast $\widehat{\mathbf{y}}_{T+h}=f(\mathcal{Z}_t)=(f_1(\mathcal{Z}_t),f_2(\mathcal{Z}_t),\ldots,f_m(\mathcal{Z}_t))^{\prime}$ using a forecasting rule  $f: \underset{t}{\otimes}\mathbf{Z}\to\mathbf{Y}$
- As always, quality of forecast can be expressed using a *loss function* $\ell():\ \mathbf{Y}\times\mathbf{Y}\to\mathbb{R}$
- In statistical setting, $\{z_t\}_{t=1}^{T+h}$ taken to come from a probability distribution $p\in\mathcal{P}$
    - Goal is to choose rule with small **risk** $E_p[\ell(\mathbf{y}_{T+h},f(\mathcal{Z}_{T}))]$
    - Empirical Risk Minimizer $\widehat{f}^{ERM}=\underset{f\in\mathbf{F}}{\arg\min}\frac{1}{T-h}\sum_{t=1}^{T}\ell(\mathbf{y}_{t+h},f(\mathcal{Z}_{t}))$ remains a fine choice 

## Multivariate Forecasting Goals

- Loss function $\ell(\mathbf{y}_{T+h},\widehat{\mathbf{y}}_{T+h})$ should express preferences over different degrees of accuracy of different parts of the forecast 
- Simplest case: **Additive loss**
    - For univariate loss functions $\{\ell_i(.,.)\}_{i=1}^{m}$ and weights $\{w_i\}_{i=1}^{m}$
    - $\ell(\mathbf{y}_{T+h},\widehat{\mathbf{y}}_{T+h})=\sum_{i=1}^{m}w_{i}\ell_{i}(y_{i,T+h},f_{i}(\mathcal{Z}_t))$
    - Weights reflect importance of each problem, and $\ell_i$ describes cost of errors in each
- When problem depends on joint performance across tasks, loss function can express this directly
    - Example: weighted square loss $\ell(\mathbf{y},\widehat{\mathbf{y}})=\sum_{i,j=1}^m w_{ij}(y_{i}-\widehat{y}_{i})(y_{j}-\widehat{y}_{j})$ 
    - Weights $w_{ij}$, restricted so $w_{ij}=w_{ji}$, reflect attention to interactions between problems
        - Additional restrictions needed to make sure loss always positive and increasing
    - If $w_{ij}=0$ for $i\neq j$, get back to additive loss 
    

## Forecasting Rules for Multivariate Forecasts

- Simplest way to build a class of multivariate forecast rules is to combine univariate classes
    - $\mathcal{F}=\times_{i=1}^{m}\mathbf{F}_i$, where $\mathcal{F}_i=\{f:\mathbf{Z}\to\mathbb{R}\}$ is a univariate class like the linear classes from before, is called **separable**
- When loss is additive AND rule class $\mathcal{F}$ **separable**, ERM reduces to solving all the problems separately
    - Weights don't matter since problems do not interact 
    - Often a reasonable choice when there is no direct relationship between the problems
- Even if problems separate, may make sense to use method combining information    
    - Stein (1956) showed an example where even with additive loss and all components of $\mathbf{y_t}$ independent, can do better than separable rules
    - Shrinkage: share info across forecasts to spread error across different components
- May use rule sets $\mathcal{F}$ where, e.g., one parameter affects forecast rule for multiple variables
    - Common in model-based macroeconomics, where theory can help specify rules
        - Called "cross-equation restrictions"
    - In cases where multiple series directly linked by accounting equations, can help to impose them
        - Called "hierarchical forecasts"
    - Example: restrict coefficients $\beta_1,\beta_2,\ldots,\beta_{m}$ in set of linear forecasts so that total size bounded
        - E.g. $\sum_{i=1}^m|\beta_i|\leq B$ or $\sum_{i=1}^m(\beta_i)^2\leq B$
  
<!-- - In case where we really only care about one component, may *still* want to make a multivariate forecast -->
<!--     - Forecasts of other variables can be an input into rule for variable of interest -->
    
  
## Vector Autoregression  

- Most common (explicitly) multivariate forecasting method by far
    - Popularized in economics by Sims (1980)
- Combines $m$ linear regression classes based on $p$ lags of all $m$ variables
    - Called a Vector Autoregression of Order $p$ or **VAR(p)**
- Forecast of $y_{it}$ depends linearly on $p$ lags of itself and of the other $m-1$ variables
$$\mathcal{F}=\{\forall i=1\ldots m,\ f_{i}(\mathcal{Y}_{t})=\beta_{i,0}+\sum_{k=1}^{m}\sum_{j=1}^{p}\beta_{i,jk}y_{k,t-j}, \beta\in\mathbb{R}^{m\times p(m+1)}\}$$
- Distinct from running $m$ separate *AR(p)* models because other variables enter in
    - All variables enter in same way in all equations, so simple to specify
- With (additive) square loss for each variable, ERM estimation can be done by running Ordinary Least Squares Regression for each outcome separately
- Also commonly estimated with weighted square loss: **Generalized Least Squares**

## Implementation and Example: GDP Forecasting Revisited

- Estimation by equation-by-equation OLS implemented as command `VAR` in library `vars`
- Input is a `ts` object containing $m$ time series and choice of order $p$
    - Options to include linear trend (`type=trend`) and seasonal dummies at frequency f  (`season=f`) in each equation 
```{r, message=FALSE,warning=FALSE}
#Libraries
library(fpp2) #Forecasting and Plotting tools
library(vars) #Vector Autoregressions
library(fredr) #Access FRED Data
library(knitr) #Use knitr to make tables
library(kableExtra) #Extra options for tables

##Obtain and transform NIPA Data (cf Lecture 08)

fredr_set_key("8782f247febb41f291821950cf9118b6") #Key I obtained for this class

#US GDP components 
GDP<-fredr(series_id = "GDP",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2019-02-12")) #Gross Domestic Product

#US GDP components from NIPA tables (cf http://www.bea.gov/national/pdf/nipaguid.pdf)
PCEC<-fredr(series_id = "PCEC",
            observation_start = as.Date("1947-01-01"),
            observation_end=as.Date("2019-02-12")) #Personal consumption expenditures
FPI<-fredr(series_id = "FPI",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2019-02-12")) #Fixed Private Investment
CBI<-fredr(series_id = "CBI",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2019-02-12")) #Change in Private Inventories
NETEXP<-fredr(series_id = "NETEXP",
              observation_start = as.Date("1947-01-01"),
              observation_end=as.Date("2019-02-12")) #Net Exports of Goods and Services
GCE<-fredr(series_id = "GCE",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2019-02-12")) #Government Consumption Expenditures and Gross Investment

#Format the series as quarterly time series objects, starting at the first date
gdp<-ts(GDP$value,frequency = 4,start=c(1947,1),names="Gross Domestic Product") 
pcec<-ts(PCEC$value,frequency = 4,start=c(1947,1),names="Consumption")
fpi<-ts(FPI$value,frequency = 4,start=c(1947,1),names="Fixed Investment")
cbi<-ts(CBI$value,frequency = 4,start=c(1947,1),names="Inventory Growth")
invest<-fpi+cbi #Private Investment
netexp<-ts(NETEXP$value,frequency = 4,start=c(1947,1),names="Net Exports")
gce<-ts(GCE$value,frequency = 4,start=c(1947,1),names="Government Spending")

#Convert to log differences to ensure stationarity and collect in frame
NIPAdata<-ts(data.frame(diff(log(gdp)),diff(log(pcec)),diff(log(invest)),diff(log(gce))),frequency = 4,start=c(1947,1),end=c(2018,3))
```
- Suppose we want to forecast GDP growth and subcomponents **C**onsumption+**I**nvestment+**G**overnment jointly
- Estimate a VAR(1) model using log differences of NIPA nominal measurements
```{r}
NIPAVAR<-VAR(NIPAdata,p=1) #VAR(1) in Y=log.diff.gdp, C=log.diff.pcec, I=log.diff.invest, G=log.diff.gce
NIPAfcst<-forecast(NIPAVAR) #Produce forecasts of all series
```
```{r,echo=FALSE}
#Collect Coefficients into a Data Frame
var1table<-data.frame(
  GDP<-NIPAVAR$varresult$diff.log.gdp..$coefficients,
  Consumption<-NIPAVAR$varresult$diff.log.pcec..$coefficients,
  Investment<-NIPAVAR$varresult$diff.log.invest..$coefficients,
  Government<-NIPAVAR$varresult$diff.log.gce..$coefficients
)
rownames(var1table)<-c("Lagged Change log Y","Lagged Change log C",
                       "Lagged Change log I","Lagged Change log G",
                       "Constant")
#Make Table of Estimated Coefficients
kable(var1table,
  col.names=c("GDP","Consumption","Investment","Government"),
  caption="Estimated VAR Coefficients") %>%
  kable_styling(bootstrap_options = "striped", font_size = 16)
```


## Forecast Results

```{r}
autoplot(NIPAfcst)+
  labs(title = "VAR(1) Forecasts of GDP and and NIPA Component Growth")
```

## Multi-period Forecasts

- Existing setup and procedures have described how to make a forecast at a single period $T+h$
- For some decision problems, may want to forecast over multiple future periods $1,2,3,\ldots,h$
    - When deciding whether to invest in long-lived durable business asset, may need to know how sales will change over lifetime of the asset
- Forecasting $(y_{T+1},y_{T+2},\ldots,y_{T+h})$ is a special case of a multivariate forecasting problem
- Simplest approach: forecast separately at each period up to horizon h
    - Apply ERM (or other rule) at each distance ahead in time
    - In case of OLS predictions, goes by name of **local projection approach**
- Local projections compute different coefficients for each period
    - Reflects different relationships between variables over time
    - Allows for flexibility especially if best predictor changes at different lengths
- Especially with many time periods, unrestricted approach has disadvantages
    - Does not exploit relationships across horizons
    - Uses many parameters or high model complexity, so is prone to overfitting
        - h times as many as 1 period
    - Can be computationally costly to run procedure $h$ times
    
    
## Alternate Approaches

- In principle, could use multivariate forecasting rule in $h$ dimensions to fit these values
    - E.g., minimize joint empirical risk over all horizons
- In most cases, including commands in `forecast` library, use a different and faster method
- **Recursive substitution** is a simple way to turn a forecast rule at horizon h=1 into a forecast for any horizon
- Especially common for autoregression forecasts, but can be applied in many cases  
- Steps
    1. Compute a single forecast rule $\widehat{f}(\mathcal{Y}_T)$ for horizon $h=1$, eg by ERM
    2. Compute prediction $\widehat{y}_{T+1}=\widehat{f}(\mathcal{Y}_T)$ and add forecast to $\mathcal{Y}_T$ to produce "augmented" data $\widehat{\mathcal{Y}}_{T+1}=(\mathcal{Y}_T,\widehat{y}_{T+1})$
    3. Use same rule to produce forecast $\widehat{y}_{T+2}=\widehat{f}(\widehat{\mathcal{Y}}_{T+1})$, and add forecast to $\widehat{\mathcal{Y}}_{T+1}$ to produce "augmented" data $\widehat{\mathcal{Y}}_{T+2}=(\widehat{\mathcal{Y}}_{T+1},\widehat{y}_{T+2})$
    4. Continue as above until horizon $h$ is reached

## Recursive Substitution for Multi-period Forecasts

- Idea is that if $\widehat{f}$ is a good forecasting rule and data distribution is stationary, it would continue to be a good forecasting rule next period, if the data to implement existed
    - This data isn't available, but if forecasts are good, the forecast might be a reasonable substitute
- The change in this sequence of forecasts as a function of horizon $h$ in response to a one unit change in $y_t$ is called the **impulse response function** of the rule
    - Useful for understanding how predictions are affected by data realizations
- Because method uses only a single fit, it is fast and does not suffer from overfitting problem to same degree as disconnected approach
    - Provides sensible way to "share" parameters across multiple forecasts 
    - Doesn't explicitly optimize multiperiod objective, but can work well **so long as** initial forecast rule is high quality

## Example: Extending an AR(2) Forecast

- Fit $\widehat{f}(\mathcal{Y}_{T})$ by OLS by minimizing square loss over AR(2) rules 
$$\widehat{\beta}=\underset{\beta_0,\beta_1,\beta_2}{\arg\min}\frac{1}{T-2}\sum_{t=2}^{T-1}(y_{t+1}-\beta_0-\beta_1y_{t}-\beta_2y_{t-1})^2$$
- Predict $\widehat{y}_{T+1}=\widehat{\beta}_0+\widehat{\beta}_1y_{T}+\widehat{\beta}_2y_{T-1}$
- Predict $\widehat{y}_{T+2}=\widehat{\beta}_0+\widehat{\beta}_1\widehat{y}_{T+1}+\widehat{\beta}_2y_{T}$ $=\widehat{\beta}_0+\widehat{\beta}_1(\widehat{\beta}_0+\widehat{\beta}_1y_{T}+\widehat{\beta}_2y_{T-1})+\widehat{\beta}_2y_{T}$
$$=(\widehat{\beta}_0+\widehat{\beta}_1\widehat{\beta}_0)+(\widehat{\beta}_1^2+\widehat{\beta}_2)y_{T}+\widehat{\beta}_1\widehat{\beta}_2y_{T-1}$$
- Predict $\widehat{y}_{T+3}=\widehat{\beta}_0+\widehat{\beta}_1\widehat{y}_{T+2}+\widehat{\beta}_2\widehat{y}_{T+1}$
$$=\widehat{\beta}_0+\widehat{\beta}_1((\widehat{\beta}_0+\widehat{\beta}_1\widehat{\beta}_0)+(\widehat{\beta}_1^2+\widehat{\beta}_2)y_{T}+\widehat{\beta}_1\widehat{\beta}_2y_{T-1})+\widehat{\beta}_2(\widehat{\beta}_0+\widehat{\beta}_1y_{T}+\widehat{\beta}_2y_{T-1})$$
$$=(\widehat{\beta}_0+\widehat{\beta}_1\widehat{\beta}_0+\widehat{\beta}_1^2\widehat{\beta}_0+\widehat{\beta}_2\widehat{\beta}_0)+(\widehat{\beta}_1^3+2\widehat{\beta}_1\widehat{\beta}_2)y_{T}+(\widehat{\beta}_1^2\widehat{\beta}_2+\widehat{\beta}_2^2)y_{T-1}$$

- At every horizon the forecast rule is an AR(2), with different coefficients
    - Acts like local projection rule, but sequence of coefficients fixed by h=1 estimate
    
## More Applications of Recursive Substitution

- Exact same procedure can be performed for multivariate forecasts, like VARs
    - `VAR` command in R (and all other VAR software I know of) produces multi-period forecasts this way
- Recursive substitution can also be used to produce univariate forecasts when data $\mathcal{Z}_t=(\mathcal{Y}_t,\mathcal{X}_t)$ used to produce forecast contains variables $\mathcal{X}_t$ other than predictor of interest 
    - Like regression forecasts from previous lecture
    - Want to predict GDP, but use multiple components to do so
- Suppose we have a forecast rule $\widehat{y}_{t+1}=\widehat{f}(\mathcal{Y}_t,\mathcal{X}_t)$
    - Cannot directly perform resursive update to obtain $\widehat{y}_{t+2}$ because $x_{t+1}$ not known
    - Solution: forecast $x_{t+1}$!
- If we extend a univariate forecast $\widehat{y}_{t+1}=\widehat{f}(\mathcal{Z}_t)$ to a multivariate forecast $(\widehat{y}_{t+1},\widehat{x}_{t+1})=(\widehat{f}(\mathcal{Z}_t),\widehat{f}_X(\mathcal{Z}_t))$, can produce updated forecasts by substitution rule
    - Adds some complexity since $x$ forecasts produced, but limits complexity of multi-period forecast  
- If original forecast is linear regression and other forecasts are also, this produces a VAR
    - VAR is natural extension of regression forecast
    - In fact, GDP equation in VAR example is exactly GDP equation in linear regression example
    
## Economic Motivation for VAR models

- Can use forecasts to replace realized values not just with a known forecast rule, but also with other equations
    - If $y_t$ related to $x_t$, replace $x_t$ with forecast to predict $y_t$
- Can be hard to think through which variables might *predict* $y_t$ in the future
    - Easier to think of variables that might be related right now
- Economic theory and intuition can suggest variables and relationships which may matter 
    - Incentives, tastes and beliefs, constraints, accounting identies, all those things you learn about in other econ classes
- If you can predict $x_t$, then through contemporaneous relationship you can predict $y_t$ (or vice-versa)
- Consider a simple example
    - Let $y_t=ax_t+u_t$ where $u_t$ is independent of $x_t$ and over time  
    - Suppose $x_t$ depends on itself and $z_{t-1}$, eg $x_t=\rho x_{t-1}+\gamma z_{t-1}+e_t$ with $e_t$ independent over time
    - Then $y_t=(a\rho) x_{t-1}+(a\gamma) z_{t-1}+(u_{t}+ae_t)$: Lagged relationship depends on strength of persistence and on current relationship
- More elaborate versions of this example inspired Sims use of VAR models in macroeconomics
    - Existing macro theory c.1980 had a lot to say about contemporaneous relationships, very little about dynamics
    - Implied relationship over time a mix of both: predictive relationships for one variable can extend to others linked indirectly
    - For accurate predictions, many variables can be relevant

## Classical Statistics Perspective

- Usual justification for autoregression forecasts is in terms of **autoregressive probability model**
$$y_{t+1}=\beta_0+\sum_{j=1}^{p}\beta_{j}y_{t-j}+e_{t+1}\text{ for all t}$$
    - Where $e_{t}$, the "residual" is independent or uncorrelated over time: stationary **white noise** 
    - And $E[e_{t+1}|\mathcal{Y}_t]=0$ for all $t$: residual is unpredictable part
- *If* data is drawn from such a probability distribution, oracle forecast for *square loss* among any *linear predictor* based on past data would be autoregressive forecast rule $\beta_0+\sum_{j=1}^{p}\beta_{j}y_{t-j}$
    - Further, oracle forecast at horizon $h$ would be the recursive update of this rule
    - If $e_{t+1}$ iid Normal random variable, this is also the Bayes forecast: even nonlinear rules can't do better
- Under some conditions on $\beta$ so that the implied distribution is stationary
    - Rule selected by ERM for any AR(d) class with $d\geq p$ gets close to oracle rule
- *If* data known to come from this model, recursive forecast not just reasonable, but optimal
- Also true for VARs and other regression forecasts with multivariate version of this model 

## Evaluating Autoregressive Forecasts

- Forecast residuals from *optimal* forecast $\widehat{e}_{t+1}=y_{t+1}-\widehat{y}_{t+1}$ are white noise in autoregressive model
- This is in fact true when forecast rule is Bayes predictor in square loss for any stationary distribution
    - Holds for VARs, regression, etc
- So if you are using a rule which is approximately the Bayes rule, residuals should be close to white noise
    - ACF of $\widehat{e}_{t+1}$ should be close to 0 everywhere
    - Use `checkresiduals` to inspect and compare various measures of distance
- Usually little reason to believe chosen family contains Bayes forecast rule
    - May not be exactly linear, or may depend on information further back than number of lags used 
    - Or even if it did that approximation is close enough that all properties of Bayes forecast hold for approximate rule
- These two issues are in conflict
    - Can always increase complexity of hypothesis class to try to get closer to Bayes forecaster
    - But doing so can worsen overfitting, making rule actually chosen further away
- Residual checks measure closeness of fit, but are unrelated to degree of overfitting
    - Generally **not** a reliable tool for finding a good forecast, as opposed to checking ex post
    - Use genuine measures of loss instead!
    
## Application to GDP Series in VAR(1) Example

```{r, class.source='fold-show'}
checkresiduals(NIPAVAR$varresult$diff.log.gdp..) #GDP Growth forecast residual diagnostic
```
- Patterns in residual plot and ACF, and formal test of white noise, suggest VAR(1) is not super close to white noise
    - Suggests patterns in data may differ somewhat from VAR(1) model: in-sample fit could be improved
    - To decide whether this or another forecast should be used, should compare loss measures 

## Summary

- Multivariate forecasting relies on same principles as single variable case
    - Can improve predictive power by including other predictors and sharing information across forecasts
- Vector Autoregressions describe most natural extension of single outcome regression models
- Multi-period forecasts are a special case of multivariate forecasts
    - Recursive substitution method provides simple and parsimonious way to extend forecasts into the future
    - Suggests using multivariate forecasts to produce better predictions even if goal is longer horizon forecasts of just one item
- Autoregressive probability model shows properties of cases in which autoregressive forecasts are particularly desirable


## Caveats

- While multivariate forecasts and models can improve quality of oracle predictions a lot
- But they can also massively increase problems with overfitting
    - VAR(p) in m variables has $m^2(p+1)$ parameters: easily huge given large number of possibly relevant predictors
    - In practice, this seriously reduces performance of these methods, makes them less attractive in practice
- Suggests seeking out methods which do more to control overfitting than ERM over linear class
    - ERM looks explicitly  only at empirical risk, not at all at generalization error
    - Popularity of VAR methods relies strongly on combination with techniques to control overfitting
- Next class: methods which take into account generalization error in a more explicit way
    - Ways to reduce complexity of class
    - Ways to choose element of class which generalizes better 

## References

- Christopher A. Sims,  "Macroeconomics and reality." *Econometrica* (1980): 1-48.
- Charles Stein, "Inadmissibility of the usual estimator for the mean of a multivariate distribution", *Proc. Third Berkeley Symp. Math. Statist. Prob.*  (1956) 1, pp. 197–206


<!-- ## Outline -->

<!-- - Overview -->
<!--     - linear function classes -->
<!-- - Reasons for multivariate forecasts -->
<!--     - Direct goal: multiple outcomes of interest -->
<!--     - Aid in univariate forecast -->
<!--     - Aid across forecasts -->
<!-- - Methods -->
<!--     - Separate ERM -->
<!--     - Joint ERM -->
<!--         - Pros v cons -->
<!-- - Canonical Procedure: Vector Autoregression   -->
<!--     - History and applications -->
<!--     - Deriving VAR from contemporaneous linear models -->
<!--     - Data Example: GDP components forecast, ctd  -->
<!-- - Multihorizon forecasts -->
<!--     - Direct Method: "Local Projection" -->
<!--     - Iterative Method (autoregressions) -->
<!--         - Pros v cons -->
<!--     - Application: multiperiod VAR forecasts -->
<!--     - Impulse Response Functions -->
<!-- - Additional practical issues -->
<!--     - Stationarity concerns?  -->
<!--         - Trends -->
<!--         - Differencing -->
<!--     - Seasonal Autoregressions? -->
<!--     - Hierarchical Forecasts? -->
<!-- - (For next class): Overfitting, and need to constrain class  -->
<!--     - Say something about residual diagnostics? (Or don't) -->




