---
title: "Structural Models"
author: "73-374 Econometrics II"
output:
  html_document:
    code_folding: show
---

## Today

- Using structural models to perform causal inference
- Represent structural relationships between variables using
    - Systems of equations
    - Causal Diagrams
- Structural models have implications for conditional independence
    - Use these to justify causal interpretation of regression


## Review of Control

- Last class, covered causal inference using *control*
- **Conditionally random assignment** is key condition
    $$Y^x\perp X|W$$
- If true, average causal effect is identified by the formula
$$E(Y^x)=\int E(Y|X=x,W=w)P(w)dw$$

- If conditional expectation function is linear, $E(Y|X=x,W=w)$ recovered by coefficient $\beta_1$ in regression
$$Y=\beta_0+\beta_1X+\gamma W+e$$
- Big question: when is conditional random assignment a reasonable assumption
    - Answer: relies on *causal model* of variables, expressed as system of equations
    
## Structural Equations

- A **structural equation** represents the effect of variables that directly cause an outcome
- Describes a model of how a variable is generated, and how it would change if other variables were changed
- $Y=f(X,U)$, $U\perp X$ is *structural* if 
$$P(Y|do(X=x))=P(f(x,U))$$
- Assuming linear structural form, i.e., for $U\perp X$, 
$$Y=\beta_0 + \beta_1X_1+\beta_2X_2+\beta_3X_3+\ldots+U$$
    - *Each* coefficient $\beta_j$ gives the direct causal effect of $X_j$ on $Y$
- A structural equation is distinct from a regression equation
    - Regression equation is what you estimate
        - Can always run, given data
    - Structural equation is model of the world
        - May or may not be same as what you run


## Structural Equation Models

- A structural equation *model* (**SEM**) describes full set of causal relationships between variables in a system
- Each **endogenous** variable $Y_1,Y_2,\ldots,Y_p$ described by structural equation 
$$Y_1=f_1(Y_2,\ldots,Y_p,U_1)$$
$$Y_2=f_2(Y_1,Y_3,\ldots,Y_p,U_2)$$
$$\vdots$$
$$Y_p=f_p(Y_1,Y_2,\ldots,Y_{p-1},U_p)$$
- Each structural equation contains only direct causes
    - Variables $Y_j$ with no effect are *excluded*
- Variables $U$ are **exogenous**
    - Not caused by any variable $Y$ in system
    
## Causal Graphs (Code)

```{r, eval=FALSE, fig.height=2}
#Library to create and analyze causal graphs
library(dagitty) 
library(ggdag) #library to plot causal graphs
yxdag<-dagify(Y~X) #create graph with arrow from X to Y
#Set position of nodes so they lie on a straight line
  coords<-list(x=c(X = 0, Y = 1),y=c(X = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(yxdag)<-coords2list(coords_df)
#Plot causal graph
ggdag(yxdag)+theme_dag_blank()+
  labs(title="X causes Y",
  subtitle="X is a parent of Y, Y is a child of X") 
```

    

## Causal Graphs

- A structural equation model can be represented as a graph
    - Variables are **nodes**: box or circle
    - Direct causal effects are (directed) **edges**: lines with arrows indicating direction
    - Nodes at source of arrow are called **parents**, nodes at ends are **children**
- Simplest example: structural equation model
    - $Y=f_1(X,U_y)$ $X=f_2(U_2)$ 
- Graph by R packages `dagitty` and `ggdag`
```{r, echo=FALSE, fig.height=2}
library(dagitty) #Library to create and analyze causal graphs
suppressWarnings(suppressMessages(library(ggdag))) #library to plot causal graphs
yxdag<-dagify(Y~X) #create graph with arrow from X to Y
#Set position of nodes so they lie on a straight line
  coords<-list(x=c(X = 0, Y = 1),y=c(X = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(yxdag)<-coords2list(coords_df)
ggdag(yxdag)+theme_dag_blank()+labs(title="X causes Y",subtitle="X is a parent of Y, Y is a child of X") #Plot causal graph
```
- $U_1,U_2$ not shown explicitly: $X$ and $Y$ are variables of interest, with exogenous random variation caused by $U_1,U_2$


## Implications of strucutural equations models

- To go from an SEM to distribution of data, *solve* the system of equations
    - Find endogenous variables $Y$ as function of only exogenous variables $U$
    - Known distribution of $U$ then gives joint distribution of $Y$
- Today, restrict interest to **acyclic** structural models
    - Causal graph does not contain path along directed edges from a variable to itself
    - Variables do not cause themselves
    - Further, assume $(U_1,\ldots,U_p)$ are *mutually independent*
- For acyclic models, solve by 
    - Start with variables with no incoming arrows
    - Substitute variable in direction of arrows for its structural equation
    - Repeat until no endogenous variable appears on lefthand side
- Resulting model is called the **reduced form**

## Example: Solving a structural model (Code) 

```{r, eval=FALSE, fig.height=3.5}
examplegraph<-dagify(Y3~Y2,Y4~Y3+Y2,Y2~Y1) #create graph
#Set position of nodes 
coords<-list(x=c(Y1 = 0, Y2 = 1, Y3 = 2, Y4 = 3),
         y=c(Y1 = 0, Y2 = 0, Y3 = -0.1, Y4 = 0)) 
coords_df<-coords2df(coords)
coordinates(examplegraph)<-coords2list(coords_df)

#Plot causal graph  
ggdag(examplegraph)+theme_dag_blank()
  +labs(title="Y1 causes Y2, 
        Y2 causes Y3, Y3 and Y2 cause Y4") 
```


## Example: Solving a structural model 
- $Y_1=f_1(U_1)$, $Y_2=f_2(Y_1,U_2)$, $Y_3=f_3(Y_2,U_3)$, $Y_4=f_4(Y_3,Y_2,U_4)$
```{r, echo=FALSE, fig.height=3.5}
examplegraph<-dagify(Y3~Y2,Y4~Y3+Y2,Y2~Y1) #create graph
#Set position of nodes 
  coords<-list(x=c(Y1 = 0, Y2 = 1, Y3 = 2, Y4 = 3),y=c(Y1 = 0, Y2 = 0, Y3 = -0.1, Y4 = 0)) 
  coords_df<-coords2df(coords)
  coordinates(examplegraph)<-coords2list(coords_df)
ggdag(examplegraph)+theme_dag_blank()+labs(title="Y1 causes Y2, Y2 causes Y3, Y3 and Y2 cause Y4") #Plot causal graph
```

- To solve, start at $Y_1$, replace in $Y_2$ to get
    - $Y_1=f_1(U_1)$, $Y_2=f_2(f_1(U_1),U_2)$
- Then substitute out for $Y_2$
    - $Y_3=f_3(f_2(f_1(U_1),U_2),U_3)$, $Y_4=f_4(Y_3,f_2(f_1(U_1),U_2),U_4)$
- Substitute out for $Y_3$ to get reduced form   
    - $Y_4=f_4(f_3(f_2(f_1(U_1),U_2),U_3),f_2(f_1(U_1),U_2),U_4)$

## Interpreting a structural equation model

- Solving an acyclic structural model gives joint distribution of endogenous variables
- What properties does the joint distribution have?
    - **Causal Markov property**: 
    - Conditional on its parents, a variable is independent of any variable that is not a *descendant*
    - $Y_k$ is a **descendant** of $Y_j$ if there is a  path along directed edges from $Y_j$ to $Y_k$
- Causal Markov Property completely defines implications of causal graph
    - Absence of an edge means conditional independence   
- In above example, applying this rule gives the following conditional indepencies
```{r}
impliedConditionalIndependencies(examplegraph)
```
    
<!--Why? Y_2 is parent of Y3, Y_1 is not a descendant of Y_3, but Y_4 is:-->    


## Every SEM tells a story

- Omitted edges imply conditional independence
    - Only remove edge if you *know* there is *no* direct effect
- Direction of edges describe direction of effect  
- **Total effect** includes indirect effect through other variables
    - In linear additive case, effect along directed path from X to Y is product of SEM coefficients
    - Total effect is sum of effects along all directed paths
- Example fitting above graph 
    - $Y4$ is wages, $Y3$ is social connections, $Y2$ is education, $Y1$ is a randomly assigned admissions decision
- Story says wages determined by social connections and education, and education affects social connections, and admission affects education but has no direct effect on other variables except through it
- Use model to make assumptions clear, and if we believe them, determine which effects we can estimate
    


<!--Add a story with plausible variables describing meaning of example SEM :-->

<!-- ## Linear Version of Above      -->

<!-- - $Y_1=U_1$, $Y_2=\beta_{0}+\beta_{1}Y_1+U_2$,  -->
<!-- - $Y_3=\gamma_{0}+\gamma_{1}Y_2+U_3$ -->
<!-- - $Y_4=\delta_{0}+\delta_{1}Y_3+\delta_{2}Y_2+U_4$ -->

<!-- - Solutions -->
<!-- $$Y_1=U_1$$ -->
<!-- $$Y_2=\beta_{0}+\beta_{1}U_1+U_2$$ -->
<!-- $$Y_3=\gamma_{0}+\gamma_{1}\beta_{0}+\gamma_{1}\beta_{1}U_1+\gamma_{1}U_2+U_3$$ -->
<!-- $$Y_4=(\delta_{0}+\delta_{1}\gamma_{0}+\delta_{1}\gamma_{1}\beta_{0}+\delta_{2}\beta_{0})+$$ -->
<!-- $$(\delta_{1}\gamma_{1}\beta_{1}+\delta_{2}\beta_{1})U_1+(\delta_{1}\gamma_{1}+\delta_{2})U_2+\delta_{1}U_3+U_4$$ -->

<!-- - Coefficient on $U_j$ in $Y_k$ represents total causal effect of independent disturbance to $Y_j$ on $Y_k$ -->

## "Do" operation and causal interventions

- The effect of a causal intervention can be explicitly calculated in an SEM using the **Do** operation
- **Do $Y_j=y$** describes action of setting a variable $Y_j$ to a specific value $y$
    - Remove equation $Y_j=f_j(...)$ and replace by $Y_j=y$
    - Solve new SEM to get distribution of any variable $P(Y_k|do(Y_j=y))$
- Equivalently, on causal graph
    - Delete all directed edges pointing in to node $Y_j$
    - New graph describes all relationships in world where $Y_j$ has been changed
- In potential outcomes notation, $Y_k^{Y_j=y}$ is distribution of $Y_k$ in the perturbed SEM   
- A "Causal effect" describes what world would be like if instead of its usual value, some variable were changed
- SEM allows calculating distribution of both observed and potential outcomes
    - Can use relationship to identify causal effects

## Example: Experiments (Code)


```{r, eval=FALSE, fig.height=1.5}
#Set position of nodes so they lie on a straight line
  coords<-list(x=c(X = 0, Y = 1),y=c(X = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(yxdag)<-coords2list(coords_df)
ggdag(yxdag)+theme_dag_blank() #Plot causal graph
```

## Example: Experiments

- In an experiment, $X$ is set (randomly) by experimenter, so has no incoming edges
    - $X=f_1(U_1)$
- Outcome of interest $Y$ has only $X$ and $U_2$, variables independent of $X$ as inputs 
    - $Y=f_2(X,U_2)$
- Causal graph is therefore

```{r, echo=FALSE, fig.height=1.5}
#Set position of nodes so they lie on a straight line
  coords<-list(x=c(X = 0, Y = 1),y=c(X = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(yxdag)<-coords2list(coords_df)
ggdag(yxdag)+theme_dag_blank() #Plot causal graph
```

- $Do(X=x)$ deletes all incoming edges to X
    - There are none, so graph is unchanged
    - Using perturbed graph $P(Y|do(X=x))=P(f_2(x,U_2))$
    - In original graph $P(Y|X=x)=P(f_2(X,U_2)|X=x)=P(f_2(x,U_2))$ by $U_1\perp U_2$
- Result: $P(Y|do(X=x))=P(Y|X=x)$
    - Exactly as derived in potential outcomes framework


<!--Structural model and corresponding graph:-->

<!--Example 2: Multivariate Experiments:-->

## Example 2: Confounding (Code 1)

```{r, eval=FALSE, fig.height=3}
confoundgraph<-dagify(Y~X+W,X~W) #create graph
#Set position of nodes 
coords<-list(x=c(X = 0, W = 1, Y = 2),
        y=c(X = 0, W = -0.1, Y = 0)) 
coords_df<-coords2df(coords)
coordinates(confoundgraph)<-coords2list(coords_df)
#Plot causal graph
ggdag(confoundgraph)+theme_dag_blank()
  labs(title="Confounding of effect of X on Y by W") 
```

## Example 2: Confounding (Code 2)

```{r, eval=FALSE, fig.height=3}
perturbedgraph<-dagify(Y~x+W) #create graph
#Set position of nodes 
coords<-list(x=c(x = 0, W = 1, Y = 2),
            y=c(x = 0, W = -0.1, Y = 0)) 
coords_df<-coords2df(coords)
coordinates(perturbedgraph)<-coords2list(coords_df)

#Plot causal graph  
ggdag(perturbedgraph)+theme_dag_blank()+
  labs(title="Perturbed Graph") 
```

## Example 2: Confounding

```{r, echo=FALSE, fig.height=3}
confoundgraph<-dagify(Y~X+W,X~W) #create graph
#Set position of nodes 
  coords<-list(x=c(X = 0, W = 1, Y = 2),y=c(X = 0, W = -0.1, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(confoundgraph)<-coords2list(coords_df)
ggdag(confoundgraph)+theme_dag_blank()+labs(title="Confounding of effect of X on Y by W") #Plot causal graph
```

- Suppose we care about effect of X on Y 
    - But W causes both X and Y
- Recover $P(Y=y|do(X=x))$ by constructing perturbed graph

```{r, echo=FALSE, fig.height=3}
perturbedgraph<-dagify(Y~x+W) #create graph
#Set position of nodes 
  coords<-list(x=c(x = 0, W = 1, Y = 2),y=c(x = 0, W = -0.1, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(perturbedgraph)<-coords2list(coords_df)
ggdag(perturbedgraph)+theme_dag_blank()+labs(title="Perturbed Graph") #Plot causal graph
```

## Recovering the adjustment formula using causal graphs

- Causal Markov property in perturbed graph implies
$$Pr(Y=y,W=w,X=x|do(X=x))=$$
$$Pr(Y=y|W=w,X=x,do(X=x))Pr(W=w|do(X=x))\cdot$$
$$Pr(X=x|do(X=x))$$
- Using that conditional laws are unchanged in perturbed graph
$$=Pr(Y=y|W=w,X=x)Pr(W=w)$$
- Object of interest $P(Y=y|do(X=x))=$
$$\int\int Pr(Y=y|W=w,X=x,do(X=x))Pr(W=w|do(X=x))\cdot$$
$$Pr(X=x|do(X=x))dwdx$$
- Substitute in to get adjustment formula: $P(Y=y|do(X=x))=$
$$\int Pr(Y=y|W=w,X=x)Pr(W=w)dw$$

## What have we learned so far?

- Introduced notation for Stuctural Equations, causal graphs
- Recover same results as in potential outcomes framework for experiments, adjustment
    - This is general: just different notation for same model
- So why bother, aside from pretty pictures?
    - Sometimes, relationship between variables takes different form
    - Rules of structural equation models can handle many new results
    - May require complicated algebra, but can be automated
- Next, show a few graphs representing alternative situations

<!--Example 4: Unobserved Confounding:-->

## Mediators

```{r, eval=FALSE, fig.height=1.5}
mediatorgraph<-dagify(Y~W,W~X) #create graph
#Set position of nodes 
coords<-list(x=c(X = 0, W = 1, Y = 2),
          y=c(X = 0, W = 0, Y = 0)) 
coords_df<-coords2df(coords)
coordinates(mediatorgraph)<-coords2list(coords_df)

#Plot causal graph
ggdag(mediatorgraph)+theme_dag_blank()+
  labs(title="Mediator structure") 
```

## Mediators

```{r, echo=FALSE, fig.height=1.5}
mediatorgraph<-dagify(Y~W,W~X) #create graph
#Set position of nodes 
  coords<-list(x=c(X = 0, W = 1, Y = 2),y=c(X = 0, W = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(mediatorgraph)<-coords2list(coords_df)
ggdag(mediatorgraph)+theme_dag_blank()+labs(title="Mediator structure") #Plot causal graph
```

- Sometimes we have access to variables $W$ called **mediators**
    - Caused by treatment $X$ and also affect outcome
    - Executioner shoots (X) Bullet hits (W) Prisoner dies (Y)
- $P(Y|do(X=x))$ asks what is outcome when $X$ happens: $=P(Y|X=x)$
- $W$ is correlated with $X$ and $Y$, but **don't** want to control for it
    - Adjustment formula gives 0 effect of $X$ on $Y$ if $W$ controlled for
    - $\int P(Y|X,W=w)P(w)dw=\int P(Y|W)P(W=w)dw=P(Y)\neq P(Y|do(X=x))$
- Conditional on being hit by a bullet, being shot at has no relationship with death
    - Changing $X$ does affect $Y$, but indirectly through $W$
- For confounders, control *removes* bias, for mediators *causes* it
    - Knowledge of causal model tells us how to interpret regression

## Colliders (Code)

```{r, eval=FALSE, fig.height=1.5}
collidergraph<-dagify(W~Y,W~X) #create graph
#Set position of nodes 
coords<-list(x=c(X = 0, W = 1, Y = 2),
               y=c(X = 0, W = 0, Y = 0)) 
coords_df<-coords2df(coords)
coordinates(collidergraph)<-coords2list(coords_df)
  
#Plot causal graph  
ggdag(collidergraph)+theme_dag_blank()+
  labs(title="Collider structure") 
```

## Colliders

```{r, echo=FALSE, fig.height=1.5}
collidergraph<-dagify(W~Y,W~X) #create graph
#Set position of nodes 
  coords<-list(x=c(X = 0, W = 1, Y = 2),y=c(X = 0, W = 0, Y = 0)) 
  coords_df<-coords2df(coords)
  coordinates(collidergraph)<-coords2list(coords_df)
ggdag(collidergraph)+theme_dag_blank()+labs(title="Collider structure") #Plot causal graph
```

- A **collider** $W$ is caused by both $X$ and $Y$
  - Grades $X$ and wealth $Y$ both help college admission $W$
- Unconditionally, $X\perp Y$ 
    - *But*, conditional on $W$, $X$ and $Y$ are dependent
    - Knowing $W$, $X$ is informative about $Y$
- Among college students, those with rich family did not need high grades to get in.
    - Observing rich family, can infer that likelihood of high grades was lower
    - True even if in full population, grades and wealth independent
- Controlling for collider in regression of $Y$ on $X$ causes bias
    - $P(Y|do(X=x))=P(Y|X=x)=P(Y)$ not equal to $\int P(Y|X,W=w)P(w)dw$
- Commonality with mediator case: $W$ is a *descendant* of $X$    

## d-separation in general graphs

- Want rules for arbitrary graphs describing when control is sufficient to recover causal effect
    - First, define rule describing what conditioning implies for independence
    
- When does conditioning on set of nodes $W$ imply for two disjoint sets of nodes $X$ $Y$ that $X\perp Y|W$?
- We say a (non-directed) path from $X_i$ to $Y_j$ is **blocked** by $W$ if it contains either
    - A collider which is *not* in $W$ and of which $W$ is *not* a descendant
    - A non-collider which *is* in $W$
- We say $X$ and $Y$ are **d-separated** (by $W$) if all paths between $X$ and $Y$ are blocked
- **Theorem** (Pearl, 2009): $X\perp Y|W$ if and only if $X$ and $Y$ are d-separated by $W$

## Backdoor Criterion

- Using d-separation, can define a complete criterion for when control recovers $P(Y|do(X=x))$
- A set of variables $W$ satisfies **Backdoor Criterion** between $X$ and $Y$ if
    - No node in $W$ is a descendant of $X$
    - $W$ blocks every path between $X$ and $Y$ that contains an edge directed into $X$ 
- **Theorem** (Pearl, 2009): If $W$ satisfies the backdoor criterion between $X$ and $Y$, the adjustment formula recovers the causal effect of $X$ on $Y$
$$P(Y|do(X=x))=\int Pr(Y|W,X=x)Pr(W)dw$$

## Backdoor criterion, intuition

- Blocking path component ensures that adjustment variables account for confounding by causes other than the cause of interest and do not introduce additional bias by inducing new correlations through colliders
- Non-descendant component avoids "bad controls" which are themselves affected by treatment
- I will not force you to recall exact definition of d-separation or backdoor criterion
- Point is that given a causal story, systematic method can recover whether control is sufficient
    - In `dagitty`, check backdoor criterion between $X$ and $Y$ by
```{r, eval=FALSE}
#Check if W satisfies criterion
isAdjustmentSet(graphname,"W",exposure="X",outcome="Y")
#Find variables that satisfy criterion, if they exist
adjustmentSets(graphname,exposure="X",outcome="Y") 
```

## Example: finding adjustment sets (Code)

```{r, eval=FALSE}
#create graph
complicatedgraph<-dagify(Y~A+C,B~X+Y,A~X,X~C,D~B) 
#Set position of nodes 
coords<-list(x=c(X = 0, A = 1, B = 1, C=1, Y=2, D=2),
    y=c(X = 0, A = 0.1, B=0, C=-0.1, Y = 0, D=0.1)) 
coords_df<-coords2df(coords)
coordinates(complicatedgraph)<-coords2list(coords_df)

#Plot causal graph
ggdag(complicatedgraph)+theme_dag_blank()+
 labs(title="Complicated Graph")

```

## Example: finding adjustment sets

```{r, echo=FALSE}
complicatedgraph<-dagify(Y~A+C,B~X+Y,A~X,X~C,D~B) #create graph
#Set position of nodes 
  coords<-list(x=c(X = 0, A = 1, B = 1, C=1, Y=2, D=2),y=c(X = 0, A = 0.1, B=0, C=-0.1, Y = 0, D=0.1)) 
  coords_df<-coords2df(coords)
  coordinates(complicatedgraph)<-coords2list(coords_df)
ggdag(complicatedgraph)+theme_dag_blank()+labs(title="Complicated Graph") #Plot causal graph

```

## Measuring effect of X on Y

- Go back to education $X$ versus wages $Y$ to get intuition
- $A$ caused by $X$, causes $Y$: e.g., occupation, experience
    - Mediators: descendant of $X$, so do not adjust for it
- $B$ caused by both $X$ and $Y$: e.g., current wealth or lifestyle
    - Colliders: descendant of $X$, so do not adjust for it
- $D$ caused by $B$ only: e.g. consequences of wealth
    - Descendants of collider: still causes bias when adjusted for
- $C$, causes $X$ and $Y$: e.g. ability, background 
    - Confounder: must condition on it 
- Backdoor criterion calculates this automatically    
```{r}
adjustmentSets(complicatedgraph,exposure="X",outcome="Y")
```

<!--## Example: School admissions

## Backdoor paths, continued

Formal definition

## Backdoor criterion

## Mediators

## "Bad Control"

## Example: Discrimination.:-->

## Summary

- Structural Equations Models describe full set of causal relationships between variables in a system
- Causal graphs represent structural equations models and the conditional independencies they imply
- Causal effects represented by "Do" operation, which asks what happens in a new graph where treatment is fixed rather than assigned by existing mechanism
- SEMs can describe situations like experiments and control, but also mediation, colliders, and more complicated structures
- Working through the implications of causal graphs yield rules to describe when adjustment by regression measures a causal effect


## References

- Short introduction
    - Kelleher, Adam. (2016) "A Technical Primer on Causality" <https://medium.com/@akelleh/a-technical-primer-on-causality-181db2575e41>
- Longer introduction
    - Nielsen, Michael. (2012) "If correlation doesn't imply causation, then what does?" <http://www.michaelnielsen.org/ddi/if-correlation-doesnt-imply-causation-then-what-does/>
- Monograph: original primary source for this material.
    - Pearl, Judea. (2009) Causality: Models, Reasoning, and Inference. 2nd ed. Cambridge UP.
- Software: R packages to analyze and display causal graphs
```{r, eval=FALSE}
    install.packages("dagitty")
    install.packages("ggdag") 
```    




