---
title: "Econometrics II"
author: "73-374 Econometrics II"
output: 
  html_document:
    code_folding: show
---
## Econometrics

>- The art of learning about economic phenomena using data
>- Neither pure economics nor statistics applied to economic data

## Goals of econometrics

- Answer scientific and policy questions
- What happens to $Y$ if we do $X$?
    - $Y=$ unemployment rate, $X=$ raise the minimum wage
    - $Y=$ our company's sales figures, $X=$ lower our prices
    - $Y=$ bus ridership, $X=$ build a new subway line
    - $Y=$ GDP growth, $X=$ raise the Fed Funds Rate

## The Econometric Approach
- An economic **model** describes behavior of economic **variables** when world is governed by a specific structure, described by a set of **parameters**
    - Parameters allow us to find causal effects.
- Goal is to use data to learn parameters.
- Separate into steps
    - **Identification**: Supposing we know exactly how certain variables behave, what would we then know about parameters 
    - **Estimation**: Find some function of the data that tells us about parameters
    - **Inference**: What values of parameters (if any) are plausibly consistent with the data? 
    
## Challenges
- Reasonable descriptions of the world often imply that the answer to our questions cannot be found in the data we have
   - `What if' questions are fundamentally asking about the behavior of something we can't see
   - Need assumptions about these unseen things, valid only in special circumstances, to make progress
- Economic data often exhibits features not well described by the most basic statistical models
    - Nonlinear relationships, dependence between observations
    - Need statistical descriptions which take these features into account
    
## This Course 

- A collection of models, methods, and examples of how to handle these challenges
- Two major topics
1. Causal inference in linear models
2. Nonlinear models and time series

## Syllabus

https://canvas.cmu.edu/


## R 
- Modern, powerful open source statistical language
- Standard at CMU, data analysts in academia and private sector
- Textbook companion at [URfIE.net](http://www.URfIE.net)
- TA will lead tutorial
- Not looking for advanced programming skills, just learn a few commands to open a data set, run some procedures.

## Example: Wage and Education Data: Code
```{r, eval=FALSE}
# Obtain access to data sets used in our textbook
library(foreign) 
# Import data set of education and wages
wage1<-read.dta(
  "http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.dta")
# Scatter plot
plot(wage1$educ,wage1$lwage, xlab = "Years of Education",
      ylab = "Log Wage", main = "Wage vs Education")
```

## Example: Wage and Education Data
```{r, echo=FALSE}
# Obtain access to data sets used in our textbook
library(foreign) 
# Import data set of education and wages
wage1<-read.dta(
  "http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.dta")
# Scatter plot
plot(wage1$educ,wage1$lwage, xlab = "Years of Education",
      ylab = "Log Wage", main = "Wage vs Education")
``` 

## Statistical Question: Do educated people earn more money?

- Data set of sample of individuals, with education and wages
- Estimation 
    - **Regression** summarizes relationship between variables in sample
- Inference
    - **Confidence Interval** for regression coefficient gives range containing population relationship with high probability
    - **Hypothesis Test** indicates whether relationship can be distinguished from a fixed value (say, 0) given sampling variation in data 
  
## Economic Question

- How much would *my* wages change if *I* stayed in school one more year?
- Answer to *statistical* question insufficient to answer this
- What kind of data can be used to answer this question?
- How do we use that data to provide an answer?
- Need an economic model of how data relates to decisions
- Fundamental statistical problems are the same
    - Estimation: how much
    - Inference: how reliable is this estimate
- Answer now depends on our understanding of how observed data relates to causal question of interest
- This affects which estimator we use, how we perform inference
- First part of class, will discuss when and how regression (and variations thereof) can answer these questions


## Statistics

- From your previous classes, you should all be able to 
    - Describe the best fitting linear relationship between education and wages 
    - Construct confidence intervals for the slope and intercept
    - Visually and numerically assess fit of the relationship
    
## Ordinary Least Squares Review   

Let ${(x_i,y_i):i=1 \ldots n}$ be a sample of measured education levels and log wages from a population of individuals.
OLS estimator solves $$(\hat{\beta_0},\hat{\beta_1})=\arg\min \sum_{i=1}^{n}(y_i-\beta_0-\beta_{1}x_i)^2$$
Giving the formulas $$\hat{\beta_1}=\frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y})}{\sum_{i=1}^{n}(x_i-\bar{x})^2}$$
$$\hat{\beta_0}=\bar{y}-\hat{\beta_1}\bar{x}$$

## Regression as statistical model
- You can run OLS on any data set
- Under some assumptions (Wooldrige Ch2), it tells us true features of the population
1. In population, $y=\beta_0+\beta_{1}x+u$
2. ${(x_i,y_i):i=1 \ldots n}$ are independent random sample of observations following 1
3. ${x_i : i=1 \ldots n}$ are not all identical
4. $E(u|x)=0$
5. $Var(u|x)=\sigma^2$ a constant $>0$


## Regression properties

- Under above assumptions, OLS estimator is 
1. Consistent: $Pr(|\hat{\beta_1}-\beta_1|>e)\rightarrow 0$ as $n\rightarrow\infty$ for any $e>0$
2. Unbiased: $E(\hat{\beta_1})=\beta_1$
3. Asymptotically normal $$Pr(\frac{\sqrt{n}(\hat{\beta_1}-\beta_1)}{\sqrt{\frac{\sigma^2} {\frac{1}{n}\sum_{i=1}^{n}(x_i-\bar{x})^2}}}<t)\rightarrow Pr(Z<t)$$ for any $t$, where $Z \sim N(0,1)$ 
- (In practice can replace $\sigma^2$ by $\frac{1}{n-2} \sum (y_i-\hat{\beta_0}-\hat{\beta_{1}}x_i)^2$ )

## Regression Results: Table (Code)
```{r, eval=FALSE, results="asis"}
# Obtain access to data sets used in our textbook
library(foreign) 
#Load library to make pretty table
suppressWarnings(suppressMessages(library(stargazer))) 
# Import data set of education and wages
wage1<-read.dta(
  "http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.dta")
# Regress log wage on years of education 
wageregoutput <- lm(formula = lwage ~ educ, data = wage1)
stargazer(wageregoutput,header=FALSE,type="html",
  omit.stat=c("adj.rsq"),font.size="tiny",
  title="Regression of Log Wage on Years of Education")
```

## Regression Results: Table
```{r, echo=FALSE, results="asis"}
# Obtain access to data sets used in our textbook
library(foreign) 
#Load library to make pretty table
suppressWarnings(suppressMessages(library(stargazer))) 
# Import data set of education and wages
wage1<-read.dta(
  "http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.dta")
# Regress log wage on years of education 
wageregoutput <- lm(formula = lwage ~ educ, data = wage1)
stargazer(wageregoutput,header=FALSE,type="html",omit.stat=c("adj.rsq"),
      font.size="tiny", title="Regression of Log Wage on Years of Education")

```

## Regression Results: Plot (Code)
```{r, eval=FALSE}
# Scatter plot with regression line
plot(wage1$educ,wage1$lwage, xlab = "Years of Education",
      ylab = "Log Wage", main = "Wage vs Education")
abline(wageregoutput,col="red")

```

## Regression Results: Plot
```{r, echo=FALSE}
# Scatter plot with regression line
plot(wage1$educ,wage1$lwage, xlab = "Years of Education",
      ylab = "Log Wage", main = "Wage vs Education")
abline(wageregoutput,col="red")

```

## Note on using R

- Regression results were implemented in **R**
- Command for linear regression

```{r, eval=FALSE}
regression<-lm(formula = lwage ~ educ)
summary(regression)
```

- Learn syntax for any command by using **?** in front of name to get help, e.g.

```{r, eval=FALSE}
?plot
```

- I will post "Code Display" version of each lecture with R code and explanatory notes for any result displayed in class
- Can also download raw code .Rmd file for each lecture

## Interpretation

- Even when assumptions are satisfied, causal question not answered by above results
- How much would wages change if I stayed in school one more year?
- Maybe education raises wages, or vice versa 
    - Or both related to some third factor (job performance?)
- Regression alone won't tell us this without extra info 
- In practice, above assumptions often dubious even as statistical descriptions
- More powerful statistical methods which better describe relationship can help
- Combination of these methods and model of economic situation *together* needed to make headway


## Causal inference strategies
- This course covers research designs which describe situations in which it is possible to extract the causal effect of education on wages 
    - Experiments
    - Control
    - Instrumental Variables
    - Fixed Effects
    - Regression Discontinuity
- We will also go over some methods for better *statistical* descriptions of the relationship.

## Nonlinear methods (code)

```{r, eval=FALSE}
#Load Libraries for nonparametric estimation
#Install "np" library 
install.packages("np", 
            repos="http://cran.us.r-project.org")
library(np))
wagebw<-npregbw(ydat=wage1$lwage,xdat=wage1$educ,
    regtype="ll",bws=c(2),
    bandwidth.compute=FALSE, ckertype="uniform")
#Pretty up Names
wagebw$xnames<-"Years of Education"
wagebw$ynames<-"Log Wage"
plot(wagebw,ylim=c(min(wage1$lwage),max(wage1$lwage)), 
    main="Wage vs Education, Linear and Nonlinear Fits")
points(wage1$educ,wage1$lwage,pch=1, col="blue")
abline(wageregoutput, col="red")

```

## Nonlinear methods

```{r, echo=FALSE}
#Load Libraries for nonparametric estimation
#If not installed, uncomment to install "np" library for nonparametrics
#install.packages("np",repos="http://cran.us.r-project.org")
suppressWarnings(suppressMessages(library(np)))
suppressWarnings(suppressMessages(wagebw<-npregbw(ydat=wage1$lwage,xdat=wage1$educ,regtype="ll",bws=c(2),bandwidth.compute=FALSE, ckertype="uniform")))
#Pretty up Names
wagebw$xnames<-"Years of Education"
wagebw$ynames<-"Log Wage"
suppressWarnings(suppressMessages(plot(wagebw,ylim=c(min(wage1$lwage),max(wage1$lwage)), main="Wage vs Education, Linear and Nonlinear Fits")))
points(wage1$educ,wage1$lwage,pch=1, col="blue")
abline(wageregoutput, col="red")
#scatter.smooth(wage1$educ,wage1$lwage, xlab = "Years of Education", ylab = "Log Wage", main = "Wage vs Education, Linear and Nonlinear Fits")

```


## Assignments and readings
- Readings
    - This week: brush up on linear models/OLS from Econometrics I/Intro to Statistical Inference
    - Wooldridge Appendix B & C Review Probability and Statistics
    - Wooldridge Ch 2-5 review linear regression, univariate and multivariate
- Assignments
    - Problem Set 1
