---
title: "Instrumental Variables continued"
author: "73-374 Econometrics II"
output:
  html_document:
    code_folding: show
---

## Instrumental Variables Continued

- Review
- Estimation and Inference
- IV in Potential outcomes framework: LATE


## Endogeneity Problem

- Suppose effect of $X$ on $Y$ can be represented as
$$Y=\beta_0+\beta_1X + u$$ 
- But with **endogeneity**
$$E[Xu]\neq 0$$


- Can arise from omitted confounders
<!-- - Additional reasons for endogeneity will be discussed next week   -->
<!--     - "Simultaneity bias": when causal graph **not** acyclic -->
<!--     - Measurement error: want  $X^{*}$, observe $X=X^{*}+e$ -->
- OLS inconsistent for $\beta_1$
    - Finds slope of (Best Linear Predictor of) Conditional Expectation Function
    - But this *isn't* what we want
    
## A Possible Solution: Instrumental Variables

- Find instrument $Z$ such that
$$Cov(Z,u)=0$$
$$Cov(Z,X)\neq 0$$
$$E[u]=0$$
- $Z$ correlated with $X$ but not residual $u$
- $Z$ affects treatment only
    - No *direct* effect on $Y$
    - No correlation with omitted confounders
- Then $\beta_1=\frac{Cov(Z,Y)}{Cov(Z,X)}$
- Estimate by **Instrumental Variables** Estimator
$$\hat{\beta}_1^{IV}=\frac{\frac{1}{n}\sum_{i=1}^{n}(Z_i-\bar{Z})(Y_i-\bar{Y})}{\frac{1}{n}\sum_{i=1}^{n}(Z_i-\bar{Z})(X_i-\bar{X})}$$
    

## IV assumptions
- Assume the following
- (IV1) Linear Model
$$Y=\beta_0+\beta_1X + u$$
- (IV2) Random sampling: $(Y_i,X_i,Z_i)$ drawn i.i.d. from population satisfying linear model assumptions
- (IV3) Relevance
$$Cov(Z,X)\neq 0$$
- (IV4(i)) Exogeneity
$$Cov(Z,u)=0$$
- (IV4(ii)) Mean 0 errors
$$E[u]=0$$
- Sometimes also assume
(IV5) Homoskedasticity
$$E[u^2|Z]=\sigma^2$$ ($\sigma^2$ a finite nonzero constant)


## IV asymptotics
- IV estimator $\hat{\beta}_1^{IV}$ is consistent for $\beta_1$ under (IV1)-(IV4(i))
- IV estimator of constant term $\hat{\beta}_0^{IV}$ consistent under (IV1)-(IV4)
- IV estimator $\hat{\beta}^{IV}$ asymptotically normal under (IV1)-(IV4)
- Standard errors of IV estimator derived in usual way using CLT under (IV1)-(IV5)
- Formula is in book, and applied by software 
    - **ivreg** in **AER** library
- Different formula if residuals are heteroskedastic 
    - (IV1)-(IV4) hold, but (IV5) doesn't 
    - Obtain from **vcovHC** command in **sandwich** library

## Example application: Estimating a demand curve
- Quantity demanded $Y$ is causally affected by price $X$
    - But also other things, including shifts in demand curve
- Want estimate of effect of price on demand
    - Sometimes firms experiment with prices, but rarely
- Instead, can find instrument $Z$ for price
    - $Z$ *relevant* if it affects price
    - $Z$ *exogenous* if independent of any shifts in demand curve 
- Need something which independently shifts supply curve
- Idea: instrument $Z$ using  
    - Random change in tax on item 
    - Random change in cost of inputs to production

## Example: Cigarette demand

- Let's estimate a demand curve for cigarettes across states (1995 data)
- Use cigarette tax increases as instrument
    - Works if tax changes determined by state fiscal needs unrelated to cigarette demand
- Price elasticity estimated by IV regression of log sales on log price, using tax difference as instrument
```{r, eval=FALSE}
ivreg(log(packs) ~ log(rprice) | tdiff, data = c1995)
```
- Decompose into reduced form: effect of tax on quantity demanded, divided by first stage: effect of taxes on prices
- OLS version of estimate gives correlation of price and quantity, but could come from supply or demand

## Example: Cigarette demand (Code)
```{r, eval=FALSE}
#Load library containing IV command 'ivreg'
library(AER)
# Load data on cigarette prices and quantities
data("CigarettesSW")
#Use real prices as X
CigarettesSW$rprice <- with(CigarettesSW, price/cpi)
#Use changes in cigarette tax as
# instrument which shifts supply curve
CigarettesSW$tdiff <- with(CigarettesSW, (taxs - tax)/cpi)
#data from different states in 1995
c1995 <- subset(CigarettesSW, year == "1995")
```

```{r, echo=FALSE}
#Load library containing IV command 'ivreg'
suppressWarnings(suppressMessages(library(AER)))
# Load data on cigarette prices and quantities
data("CigarettesSW")
#Use real prices as X
CigarettesSW$rprice <- with(CigarettesSW, price/cpi)
#Use changes in cigarette tax as supply curve shifting instrument
CigarettesSW$tdiff <- with(CigarettesSW, (taxs - tax)/cpi)
#data from different states in 1995
c1995 <- subset(CigarettesSW, year == "1995")
```

## IV estimate of demand elasticity for cigarettes (code)

```{r, eval=FALSE}
#To get IV estimate of effect of x on y using
#  z as instrument syntax is ivreg(y ~ x | z)
# Effect of log(price) on log(quantity)
# Elasticity
fm_ivreg <- ivreg(log(packs) ~ log(rprice)
                  | tdiff, data = c1995)
#Obtain (robust) standard errors
fm_ivreg.results<-coeftest(fm_ivreg,
         vcov = vcovHC(fm_ivreg, type = "HC0"))
```

## OLS estimate of demand elasticity for cigarettes (code)

```{r, eval=FALSE}
#OLS estimate of effect of log(price)
 #on log(quantity)  with no instrument
fm_ols <- lm(formula= log(packs) ~
               log(rprice), data = c1995)
#Obtain (robust) standard errors
fm_ols.results<-coeftest(fm_ols,
    vcov = vcovHC(fm_ols, type = "HC0"))
```

## Components of IV estimates of demand elasticity for cigarettes (code)

```{r, eval=FALSE}
# First stage: z on x
firststage<-lm(formula= log(rprice) ~
                 tdiff, data = c1995)
firststage.results<-coeftest(firststage,
    vcov = vcovHC(firststage, type = "HC0"))
# Reduced form
reducedform<-lm(formula= log(packs) ~
                tdiff, data = c1995)
reducedform.results<-coeftest(reducedform,
         vcov = vcovHC(reducedform, type = "HC0"))
```

## Results of IV estimates of demand elasticity for cigarettes (code)

```{r, eval=FALSE, results="asis"}
#Display results in big fancy table
library(stargazer)
stargazer(fm_ols.results,fm_ivreg.results,
          reducedform.results,firststage.results,
    type="html", 
    title="Cigarette Demand Elasticity: OLS, 
      IV, Reduced Form, First Stage",
    header=FALSE, size="tiny",
    column.labels=c("log(packs)","log(packs)",
                    "log(packs)","log(price)"),
    omit.table.layout="n")
```

## Results of IV estimates of demand elasticity for cigarettes

```{r, echo=FALSE}
#To get IV estimate of effect of x on y using
#  z as instrument syntax is ivreg(y ~ x | z)
# Effect of log(price) on log(quantity)
# Elasticity
fm_ivreg <- ivreg(log(packs) ~ log(rprice)
                  | tdiff, data = c1995)
#Obtain (robust) standard errors
fm_ivreg.results<-coeftest(fm_ivreg,
         vcov = vcovHC(fm_ivreg, type = "HC0"))
```

```{r, echo=FALSE}
#OLS estimate of effect of log(price)
 #on log(quantity)  with no instrument
fm_ols <- lm(formula= log(packs) ~
               log(rprice), data = c1995)
#Obtain (robust) standard errors
fm_ols.results<-coeftest(fm_ols,
    vcov = vcovHC(fm_ols, type = "HC0"))
```

```{r, echo=FALSE}
# First stage: z on x
firststage<-lm(formula= log(rprice) ~
                 tdiff, data = c1995)
firststage.results<-coeftest(firststage,
    vcov = vcovHC(firststage, type = "HC0"))
# Reduced form
reducedform<-lm(formula= log(packs) ~
                tdiff, data = c1995)
reducedform.results<-coeftest(reducedform,
         vcov = vcovHC(reducedform, type = "HC0"))
```

```{r, echo=FALSE, results="asis"}
#Display results in big fancy table
suppressWarnings(suppressMessages(library(stargazer)))
stargazer(fm_ols.results,fm_ivreg.results,reducedform.results,firststage.results,
    type="html", title="Cigarette Demand Elasticity: OLS, IV, Reduced Form, First Stage",
    header=FALSE, size="tiny",
    column.labels=c("log(packs)","log(packs)","log(packs)","log(price)"),
    omit.table.layout="n")
```


## Results

- OLS says 1% difference in prices associated with 1.2% lower cigarette consumption
    - Not causal effect of changing prices
- IV says 1% increase in prices (due to change in tax) leads to 1% lower cigarette consumption
- Justification: states that saw 1 % point increase in tax saw
    - 3% increase in prices (First stage)
    - 3% decrease in sales (Reduced form)
- If tax increases effectively random, can interpret both as causal
- If tax increases affect demand only through prices, ratio is effect of price on quantity
- IV elasticity $\neq$ OLS estimate: states with higher cig prices also had other characteristics which reduced cigarette sales


## Potential Outcomes Again
- Let's extend heterogeneous treatment effects model to IV case
- Recall potential outcomes model of causal effect of binary variable $X\in\{0,1\}$ on outcome $Y$
$$Y_i=Y^0_i(1-X_i)+Y^1_iX_i$$
- $Y^0_i$ is value of $Y_i$ given do $X_i=0$
- $Y^1_i$ is value of $Y_i$ given do $X_i=1$
- E.g. $X_i$ is a drug, $Y_i$ is a health outcome
- Difference of conditional averages converges to difference in conditional expectations
$$\frac{1}{n_1}\sum_{i=1}^{n}Y_iX_i-\frac{1}{n_0}\sum_{i=1}^{n}Y_i(1-X_i)\overset{p}{\to} E[Y_i|X_i=1]-E[Y_i|X_i=0]$$

    - Same as slope in regression of $X$ on $Y$ 
- If $X_i$ independent of $(Y^0_i,Y^1_i)$, e.g. because $X_i$ assigned randomly difference becomes $E[Y^1_i-Y^0_i]$
    - Average Treatment Effect (ATE)


## Potential Outcomes in IV
- In IV case, $X_i$ not random, but $Z_i$ is
    - Both binary $\{0,1\}$
- E.g., maybe you ran a controlled experiment, giving drug to one group $Z_i=1$, not to other $Z_i=0$
    - But some people forget to take the drug
- Likely that people who forget not random
    - Associated with stress, memory, physical ability, etc
- Goal is causal effect of $X_i$  on $Y_i$
- Experiment assigns $Z_i$: told to take the drug, 
    - Not $X_i$: actually take the drug
- Use $Z_i$ as instrument for $X_i$
- Random assignment of IV means $Z_i$ independent of $(Y^0_i,Y^1_i,X^0_i,X^1_i)$

## IV Estimate in Binary Z, Binary X case
- *First Stage*: Regression of $X$ on $Z$ gives $E[X_i|Z_i=1]-E[X_i|Z_i=0]$
- Since $Z$ random, this is ATE of Z on X
- *Reduced form*: Regression of $Y$ on $Z$ gives $E[Y_i|Z_i=1]-E[Y_i|Z_i=0]$
- Since $Z$ random, this is ATE of Z on Y
- IV is reduced form over first stage
$$\hat{\beta}_1^{IV}= \frac{\frac{1}{n_1}\sum_{i=1}^{n}Z_iY_i-\frac{1}{n_0}\sum_{i=1}^{n}(1-Z_i)Y_i}{\frac{1}{n_1}\sum_{i=1}^{n}Z_iX_i-\frac{1}{n_0}\sum_{i=1}^{n}(1-Z_i)X_i}$$
- Converges to 
$$\frac{E[Y_i|Z_i=1]-E[Y_i|Z_i=0]}{E[X_i|Z_i=1]-E[X_i|Z_i=0]}$$
- In constant effects case, this is $\beta$
    - But what if effects heterogeneous?

## Goal of potential outcome interpretation

- Under *constant treatment effects*, a valid IV gives ATE, independent of characteristics
- Issue is subjects induced into treatment by IV may respond differently than other subjects when made to take treatment
    - Drug more/less effective among people already healthier, and so less likely to forget
    - Students who attend a school after winning a voucher are those who expect it will help them out the most
- It turns out that, if effects differ, IV can only give you Average Treatment Effect for subpopulation
- Call this the *Local Average Treatment Effect*, **LATE**
- Still interesting if we want to give treatment just to this group
    - Next, will show who this group is

## Causal effect of Z on X
- In potential outcomes notation, $X$ given $Z$ takes value
$$X_i=X^0_i(1-Z_i)+X^1_iZ_i$$ 
- Can distinguish 4 groups
    - Always Takers: Take drug whether assigned or not
    $$X^0_i=X^1_i=1$$
    - Never Takers: Don't take drug whether assigned to or not
    $$X^0_i=X^1_i=0$$
    - Compliers: Take drug if assigned, Don't if not
    $$X^0_i=0,\ X^1_i=1$$
    - Defiers: Take drug if not assigned, don't take if assigned
    $$X^0_i=1,\ X^1_i=0$$
    
## Reduced form in potential outcomes framework
- Rewrite reduced form in terms of potential outcomes
$$E[Y_i|Z_i=1]-E[Y_i|Z_i=0]=$$ $$E[X^1_iY^1_i+(1-X^1_i)Y^0_i|Z_i=1]-E[X^0_iY^1_i+(1-X^0_i)Y^0_i|Z_i=0]$$
- Use independence: $Z_i$ independent of $(Y^0_i,Y^1_i,X^0_i,X^1_i)$
$$= E[(X^1_i-X^0_i)(Y^1_i-Y^0_i)]$$
- Decompose reduced form into effects on the 4 groups
$$= P(X_i^1=1,X_i^0=0)E[1*(Y^1_i-Y^0_i)|X_i^1=1,X_i^0=0]$$
$$+P(X_i^1=1,X_i^0=1)E[0*(Y^1_i-Y^0_i)|X_i^1=1,X_i^0=1]$$
$$+P(X_i^1=0,X_i^0=0)E[0*(Y^1_i-Y^0_i)|X_i^1=0,X_i^0=0]$$
$$+P(X_i^1=0,X_i^0=1)E[-1*(Y^1_i-Y^0_i)|X_i^1=0,X_i^0=1]$$

## Reduced form in potential outcomes framework, ctd
- Always takers and never takers unaffected by instrument
- Reduced form is weighted difference in causal effects on compliers and defiers
$$E[Y_i|Z_i=1]-E[Y_i|Z_i=0]=$$
$$= P(X_i^1=1,X_i^0=0)E[1*(Y^1_i-Y^0_i)|X_i^1=1,X_i^0=0]$$
$$+P(X_i^1=0,X_i^0=1)E[-1*(Y^1_i-Y^0_i)|X_i^1=0,X_i^0=1]$$

## Monotonicity
- In many cases, reasonable to assume no defiers
    - E.g. drug only given to treatment group, so no way for control to take it, even though treatment group may sometimes forget
    - If IV Z is something like "I give you money to do X", rational people will always be no less likely to do X when it is more profitable
- Can fail if instrument can move outcome in either direction
    - Judge A gives longer sentences to one group of criminals than Judge B, but shorter to a different group    
- If monotonicity holds, $P(X_i^1=0,X_i^0=1)=0$ and
$$E[Y_i|Z_i=1]-E[Y_i|Z_i=0]=$$
$$P(X_i^1=1,X_i^0=0)E[Y^1_i-Y^0_i|X_i^1=1,X_i^0=0]$$
- Reduced form is causal effect of $X$ for the compliers, multiplied by probability that the instrument induces compliance

## First stage in potential outcomes framework
- First stage is $E[X_i|Z=1]-E[X_i|Z_i=0]$
- Independence means this equals $E[X^1_i]-E[X^0_i]$
$$= P(X^1_i=1)-P(X^0_i=1)$$
- Apply law of total probability
$$= (P(X^1_i=1 \cap X^0_i=1)+P(X^1_i=1 \cap X^0_i=0))$$
$$-(P(X^0_i=1 \cap X^1_i=1)+P(X^0_i=1 \cap X^1_i=0))$$
$$= P(X^1_i=1 \cap X^0_i=0)-P(X^0_i=1 \cap X^1_i=0)$$
- No defiers implies
$$E[X_i|Z=1]-E[X_i|Z_i=0] = P(X^1_i=1 \cap X^0_i=0)$$
- First stage estimates proportion of compliers


## LATE
- $E[Y^1_i-Y^0_i|X_i^1=1,X_i^0=0]$ is called the LATE
    - Local Average Treatment Effect
- Recovered by $$LATE=\frac{E[Y_i|Z_i=1]-E[Y_i|Z_i=0]}{P(X_i^1=1,X_i^0=0)}$$
- Probability of compliance $P(X_i^1=1,X_i^0=0)=E[X_i|Z=1]-E[X_i|Z_i=0]$
$$LATE=\frac{E[Y_i|Z_i=1]-E[Y_i|Z_i=0]}{E[X_i|Z=1]-E[X_i|Z_i=0]}$$
- This is $\beta^{IV}$
- Under independence and monotonicity, IV estimates LATE
- With constant treatment effects, do **not** need monotonicity 
    - IV measures ATE

## Interpretation

$$\hat{\beta}^{IV}_1\overset{p}{\rightarrow}E[Y^1_i-Y^0_i|X_i^1=1,X_i^0=0]$$

- IV gives causal effect of treatment on compliers
- This is average treatment effect over a subpopulation
    - People induced to change their behavior by instrument
- Maybe not the same as population as a whole
    - People who remember to take medicine probably healthier than those who forget
    - Then LATE gives effect of drug on healthier people, not always the same as on all people
- If policy is to change $X$ for all people, IV only measures the effect if effects are the same on everyone
- If policy is to induce people to do $X$, by changing $Z$, then IV gives what we want

## Extensions
- In non-binary instrument case, IV gives weighted average of conditional causal effects over different subgroups
- See Angrist & Pischke book for many details

## Example: Charter School Effectiveness
- Policy question
    - Many states now have privately run but publicly funded schools. 
    - Do these schools improve learning outcomes? 
- In certain school districts, admission to school for some students is by lottery
    - $Z_i=1$ if admitted, $Z_i=0$ if not 
    - Independent of potential learning outcomes since random
- Not all who are admitted attend
    - $X_i=1$ if attending, $0$ otherwise
- If people who decide to attend differ from those who don't, difference in scores may provide biased measure of causal effects
- Use IV to estimate causal effect of attending a charter school on Math test scores
$$\hat{\beta}^{IV}=\frac{\text{Average score, admitted - Average score, not admitted}}{\%\text{ of admitted who attend - }\%\text{ of non-admitted who attend}}$$

## Results (from Angrist and Pischke, Table 3.1)
- Data from KIPP school in Lynn, Massachussetts, 2005-2008 
- Sample of 371 students apply to school via lottery
- 253 admitted, of whom 199 attend 
- 118 not admitted, of whom 5 attend (got in some other way)
- First stage: admitted students $199/253 - 5/118= 0.74$ percent increased likelihood of attending
- Average Math score of lottery winners: 0.00 standard deviations below state mean
- Average score for non-winners: 0.36 standard deviations below state mean
- Reduced form: 
    - Lottery winners have math scores 0.36 standard deviations higher than nonwinners. 
    - ATE says winning boosts math scores
- IV estimate: Reduced form/first stage = 0.36/0.74=0.48 standard deviations higher

## Interpretation

- Lottery boosts math scores by $.36\sigma$, and since this can only be coming from going to the school, which only 3/4 of winners did, effect of going to school is $\approx .5\sigma$ units on math tests
- Constant effects model says $0.48\sigma$ is "the" causal effect of going to charter school for this population
- If school more or less effective for some students, interpretation more limited
- Always takers: students who would go even if not admitted 
    - Probably these are people with involved parents
    - Likely to differ on education-related characteristics
- Never takers: students who won't attend even if admitted
    - Again could differ on education-related characteristics
- Defiers: students who would only go if not admitted
    - Seems reasonable to rule this out
- LATE: Causal effect of charter school for students in group who, if admitted, will choose to attend

## Conclusions

- When instrument available, IV estimator handles endogeneity
- Linear constant coefficients case: 
    - Requires **Exogeneity** and **Relevance**
    - Estimates (constant) effect
- Heterogeneous treatment effects case
    - Requires **Independence** and **Monotonicity (No Defiers)**
    - Estimates **Local Average Treatment Effect**
        - Average of Treatment Effects for *compliers* only 


## Next Class or Two

- Multivariate Instrumental Variables
    - 2SLS Estimator
- Testing IV assumptions
- More IV applications


