---
title: "Regularization"
author: "73-423 Forecasting for Economics and Business"
output:
  html_document:
    code_folding: hide
---

## Regularization

- Review of Generalization Problem and ERM
- Structural Risk Minimization Principle
    - Cross Validation
    - Penalized Estimation    
    
<!-- - Review of Generalization Problem and ERM -->
<!-- - Structural Risk Minimization Principle -->
<!-- - Estimates of Risk -->
<!--     - Cross Validation (and friends!) -->
<!-- - Penalized Estimation     -->
<!--     - Based on Number of Parameters: AIC/BIC -->
<!--     - Based on Size of Parameters: LASSO/Ridge/Elastic Net -->
<!--         - Other Penalties (weighted LASSO/group lasso/etc) -->
<!--     - Based on prediction function (Eg squared derivatives) -->
<!--         - Smoothing Splines! -->
<!--     - Duality between penalization and parameter space restriction -->
<!-- - Penalty selection -->
<!-- - Distinction between penalization and testing/model evaluation -->
<!--     - Residual Diagnostics, etc     -->
    
## Risk Minimization

- In statistical learning approach, goal is to produce rule $f\in\mathcal{F}$ with low **risk** for true distribution
    - $R_p(f)=E_p\ell(y_{T+h},f(\mathcal{Y}_{T}))$
- Empirical Risk Minimizer based on minimizing risk for empirical distribution $\widehat{R}(f)=\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_{t}))$ 
- The difference between what we want and what we achieve is the *generalization error*
    - $R_p(\widehat{f})=\widehat{R}(\widehat{f})+(R_p(\widehat{f})-\widehat{R}(\widehat{f}))$
    - Data can tell you whether first term is small: minimized by ERM
    - Theory required to ensure second term also small
- Can we do better by being more explicit about generalization error?
    - Sometimes not much: "Fundamental Theorem of Machine Learning" (Mohri et al 2018)
- Under same conditions that ERM has strong performance bound, no method can do "substantially better"
    - In particular, if $\mathcal{F}$ has "low complexity", excess risk, which is already small, cannot be made much smaller
- Does this mean we may as well always go with ERM?
    - Not really! Especially for small samples $T$ or classes with "high complexity"
    
    
## Structural Risk Minimization

- Ideal method would take into account both empirical risk and generalization error
- Impossible to do exactly since generalization error part not known 
    - But it can be approximated or bounded!
- Idea: Find a term related to generalization error, called a **penalty** $\mathcal{J}(f)$
- Use a forecasting rule, called a **Structural Risk Minimizer** which minimizes empirical risk plus penalty
$$\widehat{f}^{SRM}=\underset{f\in\mathcal{F}}{\arg\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_{t}))+\mathcal{J}(f))$$
- If penalty exactly equals generalization error, this is *exactly* the oracle estimator
- If penalized risk is a better approximation than the empirical risk, may still outperform ERM
- So, how do we find a good penalty?
    - Many proposed methods
    - Today: A selection of popular penalties and discussion of their properties
    
## Cross Validation

- We have already seen cross validation as a way to *assess* forecast quality of a forecasting method $\widehat{f}(\mathcal{Y}_{t})$
- Use a forecasting rule *constructed on a training sample* and estimates risk on different **test sample**
    - May do this for many test and training samples, then average the results
    - E.g. Risk estimated from a rolling forecast origin (cf `tsCV`)
$$\widehat{R}^{tsCV}(\widehat{f})=\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},\widehat{f}(\mathcal{Y}_{t}))$$
- Many versions exist: test and samples can be blocks, weighted in different ways, etc
- Eg: K-fold CV for ERM over independent data $\{(y_t,z_t)\}_{t=1}^{T}$
    - Split sample into K sets $\mathcal{T}_k$ of size $T/K$
    - For $k=1\ldots K$ estimate $\widehat{f}^{(-k)}=\underset{f\in\mathcal{F}}{\arg\min}\frac{1}{T-T/K}\sum_{t\notin\mathcal{T}_k}\ell(y_t,f(z_t))$
    - $\widehat{R}^{kfoldCV}=\frac{1}{T}\sum_{k=1}^{K}\sum_{t\in\mathcal{T}_k}\ell(y_t,\widehat{f}^{(-k)}(z_t))$
    - For k=T, called "Leave One Out" Cross Validation
- CV reduces bias from overfitting by using different samples for evaluation and testing    


## Cross validation over multiple hypothesis classes

- Cross validation can also be used as a way to *choose between* forecasting methods
    - Simply choose the method that minimizes the cross-validation estimate of risk
    - If methods $\widehat{f}$ are just fixed functions, for tsCV this is just ERM again
- Case where CV is useful: when we have *multiple hypothesis classes* of rules $\{\mathcal{F}_c:c\in\mathcal{C}\}$   
    - $c$ is a **tuning parameter** affecting some aspect of the procedure, often a measure of complexity
    - Eg for AR(p) classes, $\mathcal{F}_c=\{\beta_0+\sum_{j=1}^{p}\beta_{j}y_{t-j}:\ \beta\in\mathcal{R}^{p+1}\}$, tuning parameter is $p$
    - For each $c\in\mathcal{C}$, run ERM over $\mathcal{F}_c$  
$$\widehat{f}^{c}(\mathcal{Y}_t)=\underset{f\in\mathcal{F}_c}{\arg\min}\frac{1}{t-h}\sum_{s=1}^{t-h}\ell(y_{s+h},f(\mathcal{Y}_{s}))$$
    - Choose $\widehat{c}^{CV}$ by minimizing Cross-Validation risk of $\widehat{f}^{c}()$
$$\widehat{c}^{CV}=\underset{c\in\mathcal{C}}{\arg\min}\widehat{R}^{CV}(\widehat{f}^{c})$$
    - Forecast using $\widehat{f}^{\widehat{c}^{CV}}$

## Evaluating Cross Validation

- Cross validation over multiple classes not the same as ERM because $\widehat{f}^{c}$ is not a fixed function but depends on the data
- Often, classes are **ordered**: $\mathcal{F}_{c_1}\subset \mathcal{F}_{c_2}$ if $c_1<c_2$
    - Compare small classes of low complexity to big classes of high complexity: More lags, more covariates, etc
- Low complexity classes may have high empirical risk but low error on test set, vice versa for high complexity classes
- Cross validation accounts for these differences: won't always pick the highest complexity class
- Still have some approximation error, because test set distribution $\neq$ future population distribution
    - Error increases in complexity of $\mathcal{C}$, which is usually small: one or two parmeters
    - Worse when training and test samples are dependent: helped slightly by using tsCV over k-fold
- In general, provides fairly good approximation to generalization error but can be *extremely* costly to compute
    - Solve an optimization problem for each $c\in\mathcal{C}$ for each $t$ in sample
    - Can be hundreds or thousands of times as many computations as a single ERM forecast
- Variants or special cases which are less costly but provide similar or not much worse performance may be preferred

## Information Criteria

- Look for a simple measure $\mathcal{J}()$ of generalization error of rule $f$ using a measure of complexity of class $\mathcal{F}_c$, $\mathcal{J}(\mathcal{F}_c)$
- Note that in ERM, scale of risk doesn't matter: Applying increasing transform to risk gives same minimizer
- With penalization, effect on minimizer does depend on *relative* scale of criterion and penalty
- Use a *normalized* risk measure, called a *log likelihood* to ensure constant scale
    - Takes form $-2\log(p(\mathcal{Y}_T,f))$ where $p(\mathcal{Y}_T,f)$ is a probability distribution over $\mathcal{Y}_T$
    - Normalized form for square loss is $(T-h)\log(\frac{1}{T-h}\sum_{t=1}^{T-h}(y_{t+h}-f(.))^2)$
    - Cross entropy loss for binary prediction already normalized    
- A **penalized log likelihood** adds a penalty $\mathcal{J}$ to normalized risk
- Criterion value of class $c$ is the minimum value of the penalized log likelihood $\underset{f\in\mathcal{F}_c}{\min}(-2\log(p(\mathcal{Y}_T,f))+\mathcal{J}(\mathcal{F}_c))$
- Common penalties based on **number of parameters** p
    - Akaike Information Criterion, **AIC**: $\mathcal{J}(\mathcal{F}_c)=2p$
        - Also **AICc**: $\mathcal{J}(\mathcal{F}_c)=2p+\frac{2p(p+1)}{T-p-1}$
    - Bayesian Information Criterion, **BIC**: $\mathcal{J}(\mathcal{F}_c)=2p\log(T-h)$ (also called Schwarz Criterion, **SC**)
- Choose a rule which minimizes the penalized log likelihood $\widehat{f}^{\mathcal{J}}=\underset{f\in\mathcal{F}_c,c\in\mathcal{C}}{\arg\min}(-2\log(p(\mathcal{Y}_T,f))+\mathcal{J}(\mathcal{F}_c))$

## Evaluation and implementation

- AIC and BIC designed to pick a rule with low risk
- BIC has slightly larger penalty: downgrades more complex models more
    - Under some very strong conditions, AIC "close to" true risk, at least when T is large
    - BIC instead designed to pick "true model" when T is large
- For forecasting usually care about former rather than latter goal
    - For this reason, Hyndman & Athanasopolous text recommends AIC/AICc
- But be cautious: result requires large samples, and in small samples AIC can be bad measure of risk
    - For this reason, Diebold text recommends BIC, as slightly larger penalty accounts for chance of overfitting from misestimation of risk
    - AICc, closer to AIC but in between, is compromise
- Both estimates usually provide worse measure of risk than cross validation
    - Sometimes faster: $\approx T$ times fewer optimizations
        - Exception is linear models, where known formula simplifies CV calculations
- Automatically reported with `Arima`, `VAR`, `ets`, etc, or ex post by `CV`   
    - AICc also part of `auto.arima`, to choose between orders of AR
    - `VARselect` computes AIC/BIC for VAR over loss $\log\det(\sum_{t=1}^{T-h}(y_{t+h}-\widehat{y}_{t})(y_{t+h}-\widehat{y}_{t})^\prime)$  


<!-- - Criterion value of class $c$ is the value of the **penalized empirical risk** $$\underset{f\in\mathcal{F}_c}{\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_{t}))+\mathcal{J}(\mathcal{F}_c))$$ -->
<!-- - Choose $c\in\mathcal{C}$ to minimize penalized empirical risk, producing **penalized empirical risk minimizer** -->
<!-- $$\widehat{f}^{\mathcal{J}}=\underset{f\in\mathcal{F}_c,c\in\mathcal{C}}{\arg\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_{t}))+\mathcal{J}(\mathcal{F}_c))$$ -->

## Exercise

- Try out Information Criterion based evaluation approaches on Industrial Production data in a [Kaggle Notebook](https://www.kaggle.com/davidchilders/industrial-production-forecasting).

## Parameter size based penalties

- Penalization based on *number* of parameters require solving model for each parameter set $c\in\mathcal{C}$
    - Fine if only a small number: once per lag order of AR/VAR up to max lag, hard if sets $\mathcal{F}_c$ are every possible subset of p regressors: $2^p$ regressions 
- Works if best predictions come from ignoring all but a subset of predictors, and leaving rest unrestricted
- Alternative: allow all predictors to matter, but reduce total **size** of parameters by penalizing their **norm**
- Linear hypothesis classes $\mathcal{F}=\{\beta_0+\sum_{j=1}^m\beta_j z_j:\ \beta\in\mathbb{R}^{m+1}\}$, penalty is $\lambda\Vert\beta\Vert^p_p$
- Norm $\Vert\beta\Vert_p$ is a measure of "size" of coefficients, e.g.
    - L1 (LASSO) Penalty $\Vert\beta\Vert^1_1=\sum_{j=0}^d\|\beta_j\|$
    - L2 (Ridge) Penalty $\Vert\beta\Vert^2_2=\sum_{j=0}^d(\beta_j)^2$
    - Mixed L1/L2 (Elastic Net) Penalty $\Vert\beta\Vert_{1/2,a}=a\Vert\beta\Vert^1_1+(1-a)\Vert\beta\Vert^2_2$ for $a\in[0,1]$
- $\lambda\geq 0$ decides size of penalty term
    - Large value means strong restrictions: coefficients shrink a lot and class much simpler
    - Small value allows more complexity, at cost of greater overfitting risk. 0 means no restrictions: estimator is just ERM
- Penalized estimator solves $\widehat{f}^\lambda=\underset{f\in\mathcal{F}}{\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_t))+\lambda\Vert\beta\Vert_p^p)$ 


## Evaluation and Implementation

- Norm penalized estimators address overfitting by preventing large fluctuations in values of parameters
- Type of norm controls which kinds of fluctuations are more or less restricted
- L2 penalty more strongly affected by very large values, less by small values
    - All values partially shrink
- L1 penalty equally strongly affected by small and large values
    - Allows some larger coefficients, but shrinks a lot smaller ones, often to exactly 0: **sparsity**
    - Acts like penalties on # of parameters, but non-0 coefficients also slightly shrunk
- Elastic net in between, by amount controlled by $a$: small terms go to 0, but big terms also shrunk
- For a fixed penalty size $\lambda$ computation is easy: one instead of many optimizations
- Scale problem: how to choose $\lambda$?
    - Typically, find $\widehat{f}^\lambda$ for many $\lambda$ and select $\widehat{\lambda}^{CV}$ by cross validation
    - Efficient LARS algorithm makes computation for all $\lambda$ fast in linear model
- Implemented for linear hypotheses and square loss, cross-entropy, multivariate square loss, etc in `glmnet`
    - `cv.glmnet` implements cross validation parameter selection, but not time series version
    
## What kinds of model classes?

- Penalizing size of parameters can substitute for using smaller number of parameters
- Can include more predictors for same level of generalization error if penalized
    - Can use many predictors, or nonlinear functions of predictors: polynomials, splines, etc
    - Benefit is better ability to predict and get close to oracle predictor
    - Cost is that values of parameters get distorted
- How complicated can model class be?
    - Depends on number of possible relevant predictors and how big complicated a model within that class
- LASSO can handle more parameters than data points, if most are (close to) 0
    - Up to exponentially many if very sparse and large $\lambda$ chosen
- Ridge can handle arbitrary number of predictors
    - But effects of each become progressively more muted
- In practice, use LASSO with many weakly related predictors, of which only some might matter
- Use ridge with many, possibly related predictors, each of which contributes a little bit to the forecast
    - Often believable for lags, so common to use in (V)AR estimates
- Elastic net is in between: some correlated predictors, some not    
    
## Function based penalties

- Rather than penalizing parameters of predictor $f()$, you can penalize $f$ directly
- Penalized estimator solves, for $\lambda\Vert f\Vert$ some norm on $f$,
    - $\widehat{f}^\lambda=\underset{f\in\mathcal{F}}{\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_t))+\lambda\Vert f\Vert)$
- For classes $\mathcal{F}$ defined by parameters, like linear hypothesis class, little essential difference
    - (Usually) equivalent to some penalty on parameters
- But if class $\mathcal{F}$ isn't dependent on a (finite) set of parameters, eg $\mathcal{F}=\{\text{all functions of Z}\}$, can be very different
- Most prominent example: mean squared second derivative penalty $\Vert f\Vert=\int f^{\prime\prime}(z)^2dz$ 
    - Penalizes "wiggliness" regardless of particular shape: minimal for straight line 
- Amazing fact (Wahba 1991): penalized minimizer over class of all functions has known, computable form
    - Called a "Smoothing Spline": piecewise cubic polynomial between data points
    - Equivalent to ridge penalty over (T-dimensional) class of piecewise cubic polynomials
- Implemented as `smooth.spline` as well as in specialized functions in libraries `gam` or `mgcv`   
- Choose $\lambda$ by CV again: fast algorithm known for non-time series case

## Smoothing Splines: GDP growth from Lagged GDP growth

```{r,message=FALSE,warning=FALSE}
library(fpp2) #Forecasting
library(glmnet) #Lasso and Ridge
library(vars) #Vector Autoregressions
library(fredr) #Access FRED Data

##Obtain and transform NIPA Data (cf Lecture 07)

fredr_set_key("8782f247febb41f291821950cf9118b6") #Key I obtained for this class

#US GDP components 
GDP<-fredr(series_id = "GDP",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Gross Domestic Product

#US GDP components from NIPA tables (cf http://www.bea.gov/national/pdf/nipaguid.pdf)
PCEC<-fredr(series_id = "PCEC",
            observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Personal consumption expenditures
FPI<-fredr(series_id = "FPI",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Fixed Private Investment
CBI<-fredr(series_id = "CBI",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Change in Private Inventories
NETEXP<-fredr(series_id = "NETEXP",
              observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Net Exports of Goods and Services
GCE<-fredr(series_id = "GCE",
           observation_start = as.Date("1947-01-01"),
           observation_end=as.Date("2021-03-07"),
           vintage_dates = as.Date("2021-03-07")) #Government Consumption Expenditures and Gross Investment

#Format the series as quarterly time series objects, starting at the first date
gdp<-ts(GDP$value,frequency = 4,start=c(1947,1),names="Gross Domestic Product") 
pcec<-ts(PCEC$value,frequency = 4,start=c(1947,1),names="Consumption")
fpi<-ts(FPI$value,frequency = 4,start=c(1947,1),names="Fixed Investment")
cbi<-ts(CBI$value,frequency = 4,start=c(1947,1),names="Inventory Growth")
invest<-fpi+cbi #Private Investment
netexp<-ts(NETEXP$value,frequency = 4,start=c(1947,1),names="Net Exports")
gce<-ts(GCE$value,frequency = 4,start=c(1947,1),names="Government Spending")

#Convert to log differences to ensure stationarity and collect in frame
NIPAdata<-ts(data.frame(diff(log(gdp)),diff(log(pcec)),diff(log(invest)),diff(log(gce))),frequency = 4,start=c(1947,2))

# Construct lags

l1NIPA<-window(stats::lag(NIPAdata,-1),start=c(1949,2),end=c(2020,4))
l2NIPA<-window(stats::lag(NIPAdata,-2),start=c(1949,2),end=c(2020,4))
l3NIPA<-window(stats::lag(NIPAdata,-3),start=c(1949,2),end=c(2020,4))
l4NIPA<-window(stats::lag(NIPAdata,-4),start=c(1949,2),end=c(2020,4))
l5NIPA<-window(stats::lag(NIPAdata,-5),start=c(1949,2),end=c(2020,4))
l6NIPA<-window(stats::lag(NIPAdata,-6),start=c(1949,2),end=c(2020,4))
l7NIPA<-window(stats::lag(NIPAdata,-7),start=c(1949,2),end=c(2020,4))
l8NIPA<-window(stats::lag(NIPAdata,-8),start=c(1949,2),end=c(2020,4))

#convert to matrices
xvarsdf<-data.frame(l1NIPA,l2NIPA,l3NIPA,l4NIPA,l5NIPA,l6NIPA,l7NIPA,l8NIPA)
xvars<-as.matrix(xvarsdf)
NIPAwindow<-matrix(window(NIPAdata,start=c(1949,2),end=c(2020,4)),length(l1NIPA[,1]),4)

#Spline models, yeah?
library(gam) #loads smoothing spline package
library(mgcv) #A slightly different smoothing spline package
lagDlogGDP<-xvars[,1]
DlogGDP<-NIPAwindow[,1]
#Smoothing spline Regression of GDP on lagged GDP
# Default behavior fits penalty parameter by generalized Cross Validation
```

```{r,class.source = 'fold-show'}
# Predict GDP growth using smoothing spline in lagged GDP growth
gdp.spl<-smooth.spline(lagDlogGDP,DlogGDP)
splinepredictions<-predict(gdp.spl)
```
- $\lambda=$ `r gdp.spl$lambda` chosen by (non-time series) cross validation gives nonlinear fit

```{r,message=FALSE,warning=FALSE}
#Build Plot
gdpdata<-data.frame(xvars[,1],NIPAwindow[,1],splinepredictions$x,splinepredictions$y)
colnames(gdpdata)<-c("LDlGDP","DlGDP","xspline","yspline")

ggplot(gdpdata,aes(x=LDlGDP,y=DlGDP))+geom_point()+
  geom_line(aes(x=xspline,y=yspline),color="red",size=1)+
  geom_smooth(method=lm,se=FALSE,color="blue")+
  ggtitle("Smoothing Spline and Linear Fits of This vs Last Quarter's GDP Growth",
          subtitle="Spline bends to fit far off points")+
  xlab("Change in Log GDP, t-1")+ylab("Change in Log GDP, t")

# An alternate approach which uses "gam" command to fit spline
# Advantage is that it's built in to ggplot
# Disadvantage is that it uses only default options, so not customizable
# Default values here selct much higher smoothing, and give close to linear fit

 # ggplot(gdpdata,aes(x=LDlGDP,y=DlGDP))+geom_point()+
 #   ggtitle("Spline and OLS Fit of This vs  Last Quarter's GDP Growth",
 #           subtitle="There are two lines which exactly overlap")+
 #   xlab("Change in Log GDP, t-1")+ylab("Change in Log GDP, t")+
 #   geom_smooth(method=gam,se=FALSE,color="blue")+
 #   geom_smooth(method=lm,se=FALSE,color="red")
```

## Exotic penalties

- Example penalties I mentioned are just most common ones
- New and different penalties can be used to emphasize different features of the prediction or type of data used
    - Thousands in statistics/ML literature, and new ones every day
- Parameter-based penalties can emphasize different sets of parameters or features of size or shape
- Function based penalties can look at fine or large scale features, or multiple variables
    - Cases which allow finite computable solution are called (reproducing) **kernel** methods
- Shared principle: want to trade off flexible form to match patterns in observed data against size and complexity of class
    - Different prediction tasks will call for different shapes, sizes of data, so specialized penalties can help
    
## Penalties versus Bounds

- Alternately, can just run ERM over predictor class $\mathcal{F}$ with restriction that $\Vert f \Vert \leq c$ for some c
    - $\widehat{f}^{ERM,c}=\underset{f\in\mathcal{F}, \Vert f \Vert\leq c}{\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_t)))$
- E.g. only look at functions with norm of predictors bounded $\mathcal{F}_c=\{\beta_0+\sum_{j=1}^d\beta_jz_j:\ \beta\in\mathbb{R}^{d+1}, \Vert\beta\Vert_p\leq c\}$
- Can choose $c$ by cross-validation, just like with penalty
- Claim: ERM over class with bound imposed is numerically identical to some penalized empirical risk minimizer
- More formally, for any $c>0$, there is some $\widehat{\lambda}(c)$ such that
    - $\widehat{f}^{ERM,c}=\underset{f\in\mathcal{F}}{\arg\min}(\frac{1}{T-h}\sum_{t=1}^{T-h}\ell(y_{t+h},f(\mathcal{Y}_t))+\widehat{\lambda}(c)\Vert f\Vert)$
- This is application of Lagrange multiplier theorem from constrained optimization
- Slight difference is that $\widehat{\lambda}(c)$ may depend on the data, whereas it is fixed in penalized case
    - Though since $c$ and $\lambda$ usually chosen based on data anyway, even this usually not a difference
- Thus, penalization inherits properties of ERM, but with a new, restricted class of predictors
    - Explains variety of possible penalties: just different functions of the data
    

## Regularization for GDP Forecasting

```{r,message=FALSE,warning=FALSE}
library(dplyr) #Data manipulation
library(knitr) #Use knitr to make tables
library(kableExtra) #Extra options for tables
library(tibble) #More data manipulation
#Calculate information criteria
IC<-VARselect(NIPAdata,lag.max=8)

# Fit Optimal VARs by AIC, BIC
AICVAR<-VAR(NIPAdata,p=IC$selection[1])
BICVAR<-VAR(NIPAdata,p=IC$selection[3])

#Construct 1 period ahead foecasts
AICpred<-forecast(AICVAR,h=1)$forecast
BICpred<-forecast(BICVAR,h=1)$forecast

#Extract point forecasts
AICfcsts<-select(data.frame(AICpred),ends_with("Forecast"))
BICfcsts<-select(data.frame(BICpred),ends_with("Forecast"))


#Fit LASSO VAR in all variables up to 8 lags
lassovar<-glmnet(xvars,NIPAwindow,family="mgaussian")

#Show range of lambda values from unpenalized to where all variables gone
#lassovar$lambda 

#Show coefficients for large value of lambda (heavy penalization, most coefs gone)
#coef(lassovar,s=0.02)

#Show coefficients for intermediate value of lambda (medium penalization, some coefs gone)
#coef(lassovar,s=0.008)

#Show coefficients for small value of lambda (little penalization, most coefs remain)
#coef(lassovar,s=0.001)

## Fit LASSO VAR with 10-fold cross-validation
lassovarCV<-cv.glmnet(xvars,NIPAwindow,family="mgaussian")

#Lambda giving minimum 10-fold Cross Validation Error
CVlambda<-lassovarCV$lambda.min

#Baroque construction to take last period and format it for predict command
#Somehow need 2 rows in prediction value or it throws a fit 
#Presumably because matrix coerced to vector and then it thinks it can't multiply anymore
maxT<-length(NIPAwindow[,1]) #index of last period
newX<-t(as.matrix(data.frame(as.vector(t(NIPAwindow[(maxT-8):(maxT-1),])), as.vector(t(NIPAwindow[(maxT-7):maxT,])))))

#Predict 2021Q1 series
cvlassopredictions<-window(ts(predict(lassovarCV,newx = newX)[,,1],start=c(2020,4),frequency=4),start=c(2021,1))

#Fit L2 Penalized VAR in all variables up to 8 lags
ridgevar<-glmnet(xvars,NIPAwindow,alpha=0,family="mgaussian")

#Show range of lambda values from unpenalized to where all variables gone
#ridgevar$lambda 

#Show coefficients for large value of lambda (heavy penalization, coefs shrink a lot)
#coef(ridgevar,s=10)

#Show coefficients for intermediate value of lambda (medium penalization, coefs shrink a little)
#coef(ridgevar,s=1)

#Show coefficients for small value of lambda (little penalization, coefs close to OLS values)
#coef(ridgevar,s=0.01)

## Fit Ridge VAR with 10-fold cross-validation
ridgevarCV<-cv.glmnet(xvars,NIPAwindow,alpha=0,family="mgaussian")

#Lambda giving minimum 10-fold Cross Validation Error
CVlambda2<-ridgevarCV$lambda.min

#Predict 2021Q1 series
cvridgepredictions<-window(ts(predict(ridgevarCV,newx = newX)[,,1],start=c(2020,4),frequency=4),start=c(2021,1))

# #Plot Data Series
# autoplot(window(NIPAdata,start=c(2012,4)))+
#   autolayer(cvlassopredictions)+
#   autolayer(cvridgepredictions)

#  autolayer(AICpred$diff.log.gdp..$mean)+
 # autolayer(BICpred$diff.log.gdp..$mean)
```

- Let's go back to our GDP forecasting example using history of changes in log of $Y$, $C$, $I$, $G$
- Predictor class will be Vector Autoregressions in up to 8 lags of all 4 variables
    - 33 parameters per equation, 132 total
- Estimate by AIC, BIC, LASSO, and Ridge
- AIC selects `r AICVAR$p` lags, BIC selects `r BICVAR$p` lag, reflecting smaller penalty for AIC
- 10-fold cross validation selects $\lambda=$ `r CVlambda` for LASSO, $\lambda=$ `r CVlambda2` for ridge
- LASSO keeps just 4 nonzero parameters, while ridge chooses non-zero value for all 132 
    - Compare `r BICVAR$p*16+4` for BIC, `r AICVAR$p*16+4` for AIC, both ordered

```{r}
#Collect 2021Q1 Forecasts into a data frame
var1table<-data.frame(
  t(cvridgepredictions),
  t(cvlassopredictions),
  t(AICfcsts),
  t(BICfcsts)
)
rownames(var1table)<-c("GDP","Consumption","Investment","Government")
#Make Table of Estimated Coefficients
kable(var1table,
  col.names=c("Ridge","LASSO","AIC","BIC"),
  caption="2021Q1 Forecasts from Different Methods") %>%
  kable_styling(bootstrap_options = "striped", font_size = 16)
```

## Ridge Coefficients

```{r}
#Make Coefficient Table
zzz<-data.frame(as.matrix(coef(ridgevarCV)$y1),
                as.matrix(coef(ridgevarCV)$y2),
                as.matrix(coef(ridgevarCV)$y3),
                as.matrix(coef(ridgevarCV)$y4))


rownames(zzz)<-rownames(data.frame(as.matrix(coef(ridgevarCV)$y1)))
zzz %>%
mutate_all(function(x) {
    cell_spec(x, bold = T,
              color = spec_color(x))
  }) %>%
  kable(escape=F,
  col.names=c("GDP","C","I","G"),
  caption="Coefficients in Ridge Regression") %>%
  kable_styling(bootstrap_options = "striped", font_size=10)
```

## LASSO Coefficients

```{r}
#Make Coefficient Table
zzz<-data.frame(as.matrix(coef(lassovarCV)$y1),
                as.matrix(coef(lassovarCV)$y2),
                as.matrix(coef(lassovarCV)$y3),
                as.matrix(coef(lassovarCV)$y4))

rownames(zzz)<-rownames(data.frame(as.matrix(coef(lassovarCV)$y1)))
zzz %>%
mutate_all(function(x) {
    cell_spec(x, bold = T,
              color = if_else(x!=0,"red","black"))
  }) %>%
  kable(escape=F,
  col.names=c("GDP","C","I","G"),
  caption="Coefficients in LASSO Regression") %>%
  kable_styling(bootstrap_options = "striped", font_size=10)
```

## AIC Coefficients

```{r}
#Make Coefficient Table
zzz<-data.frame(AICVAR$varresult$diff.log.gdp..$coefficients,
                AICVAR$varresult$diff.log.pcec..$coefficients,
                AICVAR$varresult$diff.log.invest..$coefficients,
                AICVAR$varresult$diff.log.gce..$coefficients)

rownames(zzz)<-rownames(data.frame(AICVAR$varresult$diff.log.gdp..$coefficients))
zzz %>%
mutate_all(function(x) {
    cell_spec(x, bold = T,
              color = if_else(x!=0,"red","black"))
  }) %>%
  kable(escape=F,
  col.names=c("GDP","C","I","G"),
  caption="Coefficients in AIC Regression") %>%
  kable_styling(bootstrap_options = "striped", font_size=10)
```

## BIC Coefficients

```{r}
#Make Coefficient Table
zzz<-data.frame(BICVAR$varresult$diff.log.gdp..$coefficients,
                BICVAR$varresult$diff.log.pcec..$coefficients,
                BICVAR$varresult$diff.log.invest..$coefficients,
                BICVAR$varresult$diff.log.gce..$coefficients)

rownames(zzz)<-rownames(data.frame(BICVAR$varresult$diff.log.gdp..$coefficients))
zzz %>%
mutate_all(function(x) {
    cell_spec(x, bold = T,
              color = if_else(x!=0,"red","black"))
  }) %>%
  kable(escape=F,
  col.names=c("GDP","C","I","G"),
  caption="Coefficients in BIC Regression") %>%
  kable_styling(bootstrap_options = "striped", font_size=10)
```

<!-- ## Evaluating Autoregressive Forecasts -->

<!-- - Forecast residuals from *optimal* forecast $\widehat{e}_{t+1}=y_{t+1}-\widehat{y}_{t+1}$ are white noise in autoregressive model -->
<!-- - This is in fact true when forecast rule is Bayes predictor in square loss for any stationary distribution -->
<!--     - Holds for VARs, regression, etc -->
<!-- - So if you are using a rule which is approximately the Bayes rule, residuals should be close to white noise -->
<!--     - ACF of $\widehat{e}_{t+1}$ should be close to 0 everywhere -->
<!--     - Use `checkresiduals` to inspect and compare various measures of distance -->
<!-- - Usually little reason to believe chosen family contains Bayes forecast rule -->
<!--     - May not be exactly linear, or may depend on information further back than number of lags used  -->
<!--     - Or even if it did that approximation is close enough that all properties of Bayes forecast hold for approximate rule -->
<!-- - These two issues are in conflict -->
<!--     - Can always increase complexity of hypothesis class to try to get closer to Bayes forecaster -->
<!--     - But doing so can worsen overfitting, making rule actually chosen further away -->
<!-- - Residual checks measure closeness of fit, but are unrelated to degree of overfitting -->
<!--     - Generally **not** a reliable tool for finding a good forecast, as opposed to checking ex post -->
<!--     - Prefer measures which account for all types of error -->

<!-- ## Application to GDP Series in BIC VAR Prediction -->

<!-- ```{r} -->
<!-- checkresiduals(BICVAR$varresult$diff.log.gdp..) #GDP Growth forecast residual diagnostic -->
<!-- ``` -->
<!-- - Patterns in residual plot and ACF, and formal test of white noise, suggest VAR(1) is not super close to white noise -->
<!--     - Suggests patterns in data may differ somewhat from VAR(1) model: in-sample fit could be improved -->
<!--     - To decide whether this or another forecast should be used, should compare loss measures  -->

    
## Conclusions

- ERM does well for choosing within a simple hypothesis classes
- Minimizing penalized criterion is a sensible way to improve accounting for overfitting
    - Based on number (AIC/BIC) or size (L1/L2) of parameters or functions
- When choosing between multiple procedures, including ERM over multiple classes, can use measure of risk for each
    - Cross-validation  
- Penalization acts like restricting size of parameter class    
    - Can choose penalty by cross-validating
- Goal of regularization is to trade off in-sample and out-of-sample fit
    - Other tools exist for finding the "right" descriptive model


## References

- Mehryar Mohri, Afshin Rostamizadeh, and Ameet Talwalkar. "Foundations of Machine Learning" 2e MIT Press 2018
- Grace Wahba "Spline Models for Observational Data" (1991)
    - Overview of smoothing splines and penalization, 
    - Explains how and why one can work with all differentiable functions
    